---
title: "Local Use of dnaEPICO"
author:
    -
        name: Paul Ruiz
        affiliation:
            - Queensland University of Technology
        email: ruizpint@qut.edu.au
output:
    BiocStyle::html_document:
        self_contained: yes
        toc: true
        toc_float: true
        toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
    %\VignetteIndexEntry{2. Local Use of dnaEPICO}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL,
    message = FALSE,
    warning = FALSE
)

library(dnaEPICO)

can_run_preprocessing <- requireNamespace("minfiData", quietly = TRUE) &&
    requireNamespace("IlluminaHumanMethylation450kmanifest", quietly = TRUE) &&
    requireNamespace(
        "IlluminaHumanMethylation450kanno.ilmn12.hg19",
        quietly = TRUE
    )

can_run_sva <- requireNamespace("minfiData", quietly = TRUE)
```

# Introduction

This vignette demonstrates the local, interactive use of `dnaEPICO`. In this
mode, the main functions return structured in-memory result objects for direct
inspection in the R session. Use this route to check inputs, review
quality-control outputs, or examine result structures before larger analyses.

The examples below focus on the first three workflow stages:

1. `preprocessingMinfiEwasWater()`
2. `svaEnmix()`
3. `preprocessingPheno()`

All examples use `saveOutputs = FALSE`, which returns results in memory without
writing output files.

# Citation

The `r Biocpkg("dnaEPICO")` package uses methods from several external
packages. Because no single manuscript describes all components, the guidance
below explains how to cite `dnaEPICO` depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of
[minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd),
[ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and
[wateRmelon](https://github.com/schalkwyk/wateRmelon).

- If you use `preprocessingMinfiEwasWater()`, please cite @minfi; @minfiEPIC;
    @ENmix; @wateRmelon; @ewastools; @SWAN; @funnorm; @noob; @quantile.
- If you use `svaEnmix()`, please cite @minfi; @ENmix.


# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
This vignette assumes familiarity with a general DNA methylation workflow. For
an introduction, see this tutorial:
<https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html>.
It covers the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor
tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Users are expected to have basic familiarity with R,
Bioconductor workflows, and Illumina IDAT file structures.

For an introduction to Bioconductor, see the
[installation guide](https://www.bioconductor.org/install/).

## Probe-exclusion reference files

The lists of excluded probes depend on the Illumina methylation-array platform.

- For **Illumina HumanMethylationEPIC v2.0**, use the cross-reactive
    probe-exclusion file from
    [Peters et al. (2024)](https://doi.org/10.1186/s12864-024-10027-5).

- For **Illumina MethylationEPIC**, also known as the **850k array**, use
    the probe-exclusion resources from
    [Pidsley et al. (2016)](https://doi.org/10.1186/s13059-016-1066-1).
    The supporting files commonly used together are
    `13059_2016_1066_MOESM1_ESM.csv`,
    `13059_2016_1066_MOESM4_ESM.csv`,
    `13059_2016_1066_MOESM5_ESM.csv`, and
    `13059_2016_1066_MOESM6_ESM.csv`.

- For **Illumina HumanMethylation450k**, use the cross-reactive and
    polymorphic probe resources from
    [Chen et al. (2013)](https://doi.org/10.4161/epi.23470).

- Multiple probe-exclusion files can be supplied as a semicolon-separated
    value in `probeExclusionPath`. `dnaEPICO` reads probe IDs from each file,
    uses `probeExclusionIdColumn` when supplied, or otherwise auto-detects
    common probe-ID columns such as `ProbeID`, `TargetID`, `IlmnID`, and
    `Name`. The unique union of all probe IDs is then used to filter the
    normalised object.

- For **EPICv2**, setting `useEpicV2Manifest = TRUE` also retrieves the
    expanded Peters et al. manifest from AnnotationHub resource
    [AH116484](https://bioconductor.org/packages/EPICv2manifest). Probes
    flagged in selected manifest columns are added to the same exclusion set.
    By default, probes flagged by `CH_WGBS_evidence`, `CH_BLAT`, or
    `MissingPos` are removed, while `MismatchPos` is retained unless explicitly
    enabled.

# Installation

Install the development version of `r Biocpkg("dnaEPICO")` and its dependencies
with BiocManager:

```{r install-dnaepico, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("paulYRP/dnaEPICO")

BiocManager::valid()
```

```{r load-dnaepico, eval = FALSE, message = FALSE}
library("dnaEPICO")
```

# Step 1: preprocessingMinfiEwasWater

The preprocessing function reads a phenotype table and IDAT files, performs
quality control, normalisation, filtering, and cell-composition estimation, and
returns a list-based result object. The next chunk runs the function on a small
example input and prints the class, element names, and a preview of the
filtered phenotype table.

### Create example input files

The next chunk shows how to recreate the temporary files used in this example.
It writes a small phenotype table to a temporary directory, copies a small set
of IDAT files, and records the probe-exclusion reference used during probe
filtering.

```{r preprocessing-local-inputs, eval = can_run_preprocessing}
preprocessing_inputs <- dnaEPICO:::exampleMinfiIdatInputsDnaEpico(n = 6L)

names(preprocessing_inputs)
preprocessing_inputs$tempDir
basename(preprocessing_inputs$phenoFile)
head(
    basename(list.files(
        preprocessing_inputs$idatFolder,
        full.names = TRUE
    )),
    4
)
basename(preprocessing_inputs$probeExclusionPath)
```

`preprocessing_inputs` is a small list returned by the internal example helper.
`names(preprocessing_inputs)` shows its main elements: `tempDir` is the
temporary working directory, `idatFolder` contains the copied example IDAT
files, `phenoFile` is the phenotype table used by the function, `targets` is the
same phenotype information already loaded into memory, `arrayType` and
`annotationVersion` describe the array platform, and `probeExclusionPath`
points to the probe-exclusion reference used later during filtering.

```{r preprocessing-local, eval = can_run_preprocessing}
preprocessResult <- preprocessingMinfiEwasWater(
    phenoFile = preprocessing_inputs$phenoFile,
    idatFolder = preprocessing_inputs$idatFolder,
    outputLogs = file.path(preprocessing_inputs$tempDir, "logs"),
    nSamples = 6,
    SampleID = "Sample_Name",
    arrayType = preprocessing_inputs$arrayType,
    annotationVersion = preprocessing_inputs$annotationVersion,
    scriptLabel = "preprocessingMinfiEwasWater",
    baseDataFolder = file.path(preprocessing_inputs$tempDir, "rData"),
    figureBaseDir = file.path(preprocessing_inputs$tempDir, "figures"),
    detPThreshold = 1,
    normMethods = "quantile",
    sexColumn = "Sex",
    removeSexMismatch = FALSE,
    pvalThreshold = 1,
    chrToRemove = "",
    snpsToRemove = "SBE",
    mafThreshold = 1,
    probeExclusionPath = preprocessing_inputs$probeExclusionPath,
    plotGroupVar = "Sex",
    lcRef = "saliva",
    phenoOrder = "Sample_Name;Sex;Basename;Sentrix_ID;Sentrix_Position",
    lcPhenoDir = preprocessing_inputs$tempDir,
    display = FALSE,
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = FALSE
)

class(preprocessResult)
names(preprocessResult)
head(preprocessResult$targets[, c(
    "Sample_Name",
    "Sex",
    "Sentrix_ID",
    "Sentrix_Position"
)])
```

The returned object has class `dnaEPICO_preprocessingMinfiEwasWater`.
`names(preprocessResult)` lists the main elements returned by the function. The
`targets` element is the filtered phenotype table aligned to the retained
samples. The remaining elements are nested result objects for the major
preprocessing stages, including the `RGSet`, the methylation metrics, and the
cell-composition results.

Arguments:

- `phenoFile` and `idatFolder` define the phenotype table and the directory of
    raw IDAT pairs.
- `nSamples = 6` keeps the example small enough for interactive use.
- `SampleID = "Sample_Name"` identifies the column used to align phenotype data
    and methylation objects.
- `arrayType` and `annotationVersion` must match the array platform being
    processed.
- `detPThreshold`, `normMethods`, `pvalThreshold`, `chrToRemove`,
    `snpsToRemove`, and `mafThreshold` control quality filtering and probe
    filtering.
- `probeExclusionPath` supplies the probe-exclusion reference used during
    filtering.
- `sexColumn` identifies reported sex for comparison with methylation-predicted
    sex. `removeSexMismatch = FALSE` retains and reports confirmed mismatches;
    `TRUE` removes samples only when both values are known and disagree. Missing
    or unknown sex values are retained.
- `plotGroupVar`, `lcRef`, and `phenoOrder` control quality-control grouping,
    cell-type estimation, and the structure of the returned phenotype table.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
    `saveOutputs = FALSE` keep the example quiet, in memory, and free of file
    outputs.

The next chunk inspects three important elements: the filtered `RGSet`, the
metric matrices, and the phenotype table augmented with cell-composition
estimates.

```{r preprocessing-local-inspect, eval = can_run_preprocessing}
class(preprocessResult$RGSet)
names(preprocessResult$metricsData)
head(preprocessResult$lcData$phenoLC[, 1:6])
```

In local use, these are the most useful objects to inspect first.
`preprocessResult$RGSet` is the core methylation object. `metricsData`
contains the beta-value, M-value, and copy-number matrices used in downstream
steps. `lcData$phenoLC` is the phenotype table augmented with estimated cell
composition and is the object passed forward to later workflow stages.

# Step 2: svaEnmix

The SVA function loads a phenotype table and a saved `RGSet`, estimates
surrogate variables from ENmix control probes, and returns both the surrogate
variables and the merged phenotype table.

### Create example input files

The next chunk shows how to recreate the temporary files consumed by
`svaEnmix()`: a phenotype CSV file and a saved `RGSet` object in `.RData`
format.

```{r sva-local-inputs, eval = can_run_sva, message = FALSE, warning = FALSE}
sva_inputs <- dnaEPICO:::exampleMinfiBaseDataDnaEpico()
sva_temp_dir <- tempdir()
sva_targets_file <- file.path(sva_temp_dir, "sva_targets.csv")
sva_rgset_file <- file.path(sva_temp_dir, "sva_RGSet.RData")

names(sva_inputs)
class(sva_inputs$RGSet)
utils::write.csv(sva_inputs$targets, sva_targets_file, row.names = FALSE)
RGSet <- sva_inputs$RGSet
save(RGSet, file = sva_rgset_file)

basename(c(sva_targets_file, sva_rgset_file))
file.exists(c(sva_targets_file, sva_rgset_file))
```

`sva_inputs` is another helper list used to make the vignette self-contained.
`names(sva_inputs)` shows that it provides the in-memory `RGSet`, the matching
`targets` table, and a `probeExclusionPath` reference. In this step, the
function
needs file inputs rather than the in-memory objects, so the chunk writes
`targets` to `sva_targets.csv` and saves the `RGSet` object to
`sva_RGSet.RData`.

```{r sva-local, eval = can_run_sva, message = FALSE, warning = FALSE}
svaResult <- svaEnmix(
    phenoFile = sva_targets_file,
    rgsetData = sva_rgset_file,
    outputLogs = file.path(tempdir(), "logs"),
    SampleID = "Sample_Name",
    arrayType = "IlluminaHumanMethylation450k",
    annotationVersion = "ilmn12.hg19",
    SentrixIDColumn = "Sentrix_ID",
    SentrixPositionColumn = "Sentrix_Position",
    ctrlSvaPercVar = 0.90,
    ctrlSvaFlag = 1,
    scriptLabel = "svaEnmix",
    display = FALSE,
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = FALSE
)

class(svaResult)
names(svaResult)
dim(svaResult$svaData$sva)
head(svaResult$mergedPheno[, 1:min(6, ncol(svaResult$mergedPheno))])
```

The returned object has class `dnaEPICO_svaEnmix`, and `names(svaResult)`
shows the main SVA components returned by the function. The `svaData$sva`
matrix contains one row per sample and one column per surrogate variable. The
`mergedPheno` table is often the most convenient object to inspect because it
combines the original phenotype data and the new SVA columns in one place.

Arguments:

- `phenoFile` points to the phenotype table that already includes the cell
    composition estimates from preprocessing.
- `rgsetData` points to the saved `RGSet` used to derive ENmix control-probe
    surrogate variables.
- `SampleID`, `SentrixIDColumn`, and `SentrixPositionColumn` identify the
    sample and array-position columns used in the SVA analysis.
- `ctrlSvaPercVar = 0.90` asks the procedure to keep enough control-derived
    surrogate variables to explain 90% of the control-probe variance.
- `ctrlSvaFlag = 1` enables the control-based SVA workflow.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
    `saveOutputs = FALSE` keep the example interactive and in-memory.

The association analysis between the surrogate variables and array-position
metadata is stored separately in `analysisData`.

```{r sva-local-inspect, eval = can_run_sva, message = FALSE, warning = FALSE}
class(svaResult$analysisData)
names(svaResult$analysisData)
```

This object contains the fitted models and ANOVA summaries used to assess how
the surrogate variables relate to Sentrix chip and position effects.

# Step 3: preprocessingPheno

The phenotype-preparation function aligns the phenotype table with the beta,
M-value, and copy-number matrices; splits the data by timepoint; constructs a
combined longitudinal object; and prepares Clock Foundation export tables.

### Create example input files

The next chunk shows how to recreate the temporary phenotype and matrix files
used in this example. The helper writes a phenotype table plus aligned beta,
M-value, and copy-number objects to a temporary directory.

```{r preprocessingPheno-local-inputs, message = FALSE, warning = FALSE}
pheno_inputs <- dnaEPICO:::examplePreprocessingPhenoStateDnaEpico()

names(pheno_inputs)
pheno_inputs$tempDir
basename(c(
    pheno_inputs$phenoPath,
    pheno_inputs$betaPath,
    pheno_inputs$mPath,
    pheno_inputs$cnPath
))
```

`pheno_inputs` is a list that bundles both the temporary files and the
corresponding in-memory objects used in this stage. `names(pheno_inputs)` shows
that it contains the temporary directory, the phenotype table, the three saved
matrix files (`phenoPath`, `betaPath`, `mPath`, and `cnPath`), and precomputed
helper objects such as `metricsData`, `timepointData`, `combinedData`, and
`clockFoundation`.

```{r preprocessingPheno-local, message = FALSE, warning = FALSE}
phenoResult <- preprocessingPheno(
    phenoFile = pheno_inputs$phenoPath,
    betaPath = pheno_inputs$betaPath,
    mPath = pheno_inputs$mPath,
    cnPath = pheno_inputs$cnPath,
    SampleID = "Sample_Name",
    timeVar = "Timepoint",
    timepoints = "1,2",
    combineTimepoints = "1,2",
    outputPheno = file.path(pheno_inputs$tempDir, "data", "preprocessingPheno"),
    outputRData = file.path(
        pheno_inputs$tempDir,
        "rData",
        "preprocessingPheno",
        "metrics"
    ),
    outputRDataMerge = file.path(
        pheno_inputs$tempDir,
        "rData",
        "preprocessingPheno",
        "mergeData"
    ),
    sexColumn = "Sex",
    outputLogs = file.path(pheno_inputs$tempDir, "logs"),
    outputDir = file.path(pheno_inputs$tempDir, "clockFoundation"),
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = FALSE
)

class(phenoResult)
names(phenoResult)
names(phenoResult$combinedData)
head(phenoResult$combinedData$phenoMethylation[, 1:6])
```

The returned object has class `dnaEPICO_preprocessingPheno`, and
`names(phenoResult)` shows the main returned components. The
`combinedData` element groups the combined longitudinal objects, and
`phenoResult$combinedData$phenoMethylation` is the key downstream table for
modeling. By default it combines phenotype columns and CpG beta values in a
single data frame that can be passed to the GLM or LME functions. Set
`methylationScale = "m"` for M-values or `"cn"` for copy-number values.

Arguments:

- `phenoFile`, `betaPath`, `mPath`, and `cnPath` define the phenotype table and
    the three methylation matrices that are aligned in this step.
- `SampleID = "Sample_Name"` is the key used to match samples across the loaded
    objects.
- `timeVar = "Timepoint"`, `timepoints = "1,2"`, and
    `combineTimepoints = "1,2"` define the separate exports and the combined
    longitudinal object.
- `outputPheno`, `outputRData`, `outputRDataMerge`, and `outputDir` still need
    to be specified even when `saveOutputs = FALSE`, because they define the
    canonical output structure for the function.
- `sexColumn = "Sex"` identifies the sex variable used by the helper functions.
- `verbose = FALSE`, `logs = FALSE`, and `saveOutputs = FALSE` keep the example
    focused on the returned object rather than file writing.

The Clock Foundation export objects are returned in the same result, which
makes it possible to inspect the export-ready tables before deciding whether
they should be used in a reporting or export workflow.

```{r preprocessingPheno-local-inspect, message = FALSE, warning = FALSE}
names(phenoResult$clockFoundation)
head(phenoResult$clockFoundation$phenoCF)
```

The `clockFoundation` element contains the export-ready objects, and
`phenoResult$clockFoundation$phenoCF` is the phenotype table prepared for
[Clock Foundation](https://dnamage.clockfoundation.org/user) use.

## Variables used in GLM and LME models

The phenotype-methylation tables returned by `preprocessingPheno()` can be used
by `methylationGLM()` or `methylationLME()`. `factorVars` identifies categorical
fixed-effect variables, while `scaleVars` identifies numeric fixed-effect
variables to centre and divide by their sample standard deviations before model
fitting. For example, `factorVars = "Sex,TreatmentGroup"` treats those variables
as categorical and `scaleVars = "Age,PC1"` standardises `Age` and `PC1`.

Variables in `scaleVars` must be numeric, contain at least two finite
observations, and have non-zero sample variance. A variable cannot appear in
both `factorVars` and `scaleVars`. Applied centres and sample standard
deviations are stored in the annotated workbook metadata.

The preprocessed beta, M-value, or copy-number columns are passed unchanged to
the selected model engine. Their observed finite minima and maxima are reported
during preprocessing. GLM, lmerTest/lme4, and nlme messages, warnings, or errors
are stored in one `<Phenotype>_Model.Message` column. A CpG is included in the
annotated workbook when at least one coefficient or omnibus p-value is returned;
workbook metadata counts model attempts with and without p-values.

Longitudinal lmerTest/lme4 analyses use a participant-level random intercept,
`(1 | personVar)`. Set `omnibusTest = TRUE` to obtain one joint F test per CpG.
When `interactionTerm` is supplied, the joint null covers all estimable
phenotype-by-interaction coefficients; otherwise, it covers the phenotype main
effect. `omnibusDdf = "Satterthwaite"` is the computationally efficient default,
while `"Kenward-Roger"` is available when `pbkrtest` is installed. The nlme
backend does not currently support this option.

# Summary

In interactive use, inspect the returned object after each step. Useful checks
include:

- `class(result)` to confirm which function produced the object
- `names(result)` to see the main return elements
- `head()` on the phenotype-like tables
- `class()` or `dim()` on the methylation objects and matrices

The next vignette shows how to use `saveOutputs = TRUE` and the Makefile
workflow. The file-based vignette also describes the regression and
mixed-effects model structures used downstream, including how phenotype-specific
PRS terms are added through `prsMap`.

# Basics

Date the vignette was generated.

```{r vignette-date, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r vignette-wallclock, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r vignette-session-info, echo = FALSE}
sessionInfo()
```

## Asking for help

Use the [Bioconductor support site](https://support.bioconductor.org/) for
package questions. Add the `dnaEPICO` tag, review
[previous posts](https://support.bioconductor.org/tag/dnaEPICO/), and include a
small reproducible example with session information.

# References
