---
title: "File-Based and Pipeline Use of dnaEPICO"
author:
  - name: Paul Ruiz
    affiliation:
      - Queensland University of Technology
    email: ruizpint@qut.edu.au
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
  %\VignetteIndexEntry{File-Based and Pipeline Use of dnaEPICO}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL
)

library(dnaEPICO)

pheno_inputs <- dnaEPICO:::examplePreprocessingPhenoStateDnaEpico()
glm_inputs <- dnaEPICO:::exampleMethylationGLMStateDnaEpico()

can_run_sva <- requireNamespace("minfiData", quietly = TRUE)
can_run_glm <- requireNamespace(
  "IlluminaHumanMethylation450kanno.ilmn12.hg19",
  quietly = TRUE
)
sva_targets_file <- NULL
sva_rgset_file <- NULL
if (isTRUE(can_run_sva)) {
  sva_inputs <- dnaEPICO:::exampleMinfiBaseDataDnaEpico()
  sva_temp_dir <- tempdir()
  sva_targets_file <- file.path(sva_temp_dir, "pipeline_sva_targets.csv")
  sva_rgset_file <- file.path(sva_temp_dir, "pipeline_sva_RGSet.RData")
  utils::write.csv(sva_inputs$targets, sva_targets_file, row.names = FALSE)
  RGSet <- sva_inputs$RGSet
  save(RGSet, file = sva_rgset_file)
}
```

# Introduction

This vignette describes the file-based route in `dnaEPICO`. In this mode, the
same wrapper functions still return structured result objects, but they also
write files to disk when `saveOutputs = TRUE`. This route is appropriate when
you want a reproducible folder structure, Makefile-driven execution, or a
pipeline that can be resumed on a local workstation or an HPC system.

# Citation

The `r Biocpkg("dnaEPICO")` package uses methods from several external
packages. Because no single manuscript describes all components, the guidance
below explains how to cite `dnaEPICO` depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of
[minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd),
[ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and
[wateRmelon](https://github.com/schalkwyk/wateRmelon/blob/master/vignettes/references.bib).

- If you use `make f3 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
  @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools.
- If you use `make f4 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
  @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2.
- If you use `make f3lme MODEL=model1`, please cite @minfi; @minfiEPIC;
  @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @lme4.
- If you use `make all MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
  @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2; @lme4.

# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
To read this vignette, we assume that you are already familiar with the general
DNA methylation workflow. If not, we recommend first reading
[this tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html),
which provides a practical introduction to the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor
tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Downstream statistical modelling relies on base R
and CRAN frameworks, including generalised linear models and linear
mixed-effects models. Users are expected to have basic familiarity with R,
command-line execution, and Illumina IDAT file structures.

If you are asking yourself the question "Where do I start using
Bioconductor?", you might be interested in
[this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

# Installation

The `r Biocpkg("dnaEPICO")` package depends on `r Biocpkg("minfi")`,
`r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`, among others. Install
dependencies with BiocManager:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("dnaEPICO")

BiocManager::valid()
```

```{r, eval = FALSE, message = FALSE}
library("dnaEPICO")
```

# Extract the example Makefile

The package includes a template Makefile that can be copied into a project
directory and adapted to local paths and model definitions.

```{r makefile-example}
makefilePath <- extractMake(destDir = tempdir(), overwrite = TRUE)
basename(makefilePath)
```

After extraction, the file should be edited so that the project-specific paths,
model labels, and cluster settings match the target analysis environment.

Arguments used in this example:

- `destDir = tempdir()` writes the template to a temporary directory so the
  example does not modify the working tree.
- `overwrite = TRUE` allows the template to be replaced if it already exists in
  that temporary location.

# Write SVA outputs to disk

The next example shows `svaEnmix()` in file-writing mode. The function still
returns a structured result object, but the `savedFiles` element now records the
main output files written to disk.

```{r sva-pipeline, eval = can_run_sva, message = FALSE, warning = FALSE}
svaPipelineResult <- svaEnmix(
  phenoFile = sva_targets_file,
  rgsetData = sva_rgset_file,
  outputLogs = file.path(tempdir(), "logs"),
  SampleID = "Sample_Name",
  arrayType = "IlluminaHumanMethylation450k",
  annotationVersion = "ilmn12.hg19",
  SentrixIDColumn = "Sentrix_ID",
  SentrixPositionColumn = "Sentrix_Position",
  ctrlSvaPercVar = 0.90,
  ctrlSvaFlag = 1,
  scriptLabel = "svaEnmix",
  dataBaseDir = file.path(tempdir(), "data"),
  rBaseDir = file.path(tempdir(), "rData"),
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = TRUE
)

class(svaPipelineResult$savedFiles)
names(svaPipelineResult$savedFiles)
basename(unlist(svaPipelineResult$savedFiles, use.names = FALSE))
```

The file outputs from this step include the surrogate-variable matrix, the
phenotype table with appended SVA columns, and the serialized SVA object used in
later stages.

Arguments used in this example:

- `phenoFile` and `rgsetData` identify the phenotype table and saved `RGSet`
  produced by the earlier preprocessing step.
- `SampleID`, `SentrixIDColumn`, and `SentrixPositionColumn` specify how
  samples and array positions are identified in the phenotype table.
- `ctrlSvaPercVar = 0.90` keeps enough control-derived surrogate variables to
  explain 90% of the control-probe variance.
- `ctrlSvaFlag = 1` enables the ENmix control-based SVA workflow.
- `dataBaseDir` and `rBaseDir` define where file outputs are written when
  `saveOutputs = TRUE`.
- `display = FALSE`, `verbose = FALSE`, and `logs = FALSE` keep the example
  quiet while still demonstrating file creation.

# Write preprocessingPheno outputs to disk

`preprocessingPheno()` is the step that usually prepares the largest number of
pipeline-ready files because it writes timepoint-specific tables, combined
longitudinal tables, and the Clock Foundation export inputs.

```{r preprocessingPheno-pipeline, message = FALSE, warning = FALSE}
phenoPipelineResult <- preprocessingPheno(
  phenoFile = pheno_inputs$phenoPath,
  betaPath = pheno_inputs$betaPath,
  mPath = pheno_inputs$mPath,
  cnPath = pheno_inputs$cnPath,
  SampleID = "Sample_Name",
  timeVar = "Timepoint",
  timepoints = "1,2",
  combineTimepoints = "1,2",
  outputPheno = file.path(pheno_inputs$tempDir, "data", "preprocessingPheno"),
  outputRData = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "metrics"
  ),
  outputRDataMerge = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "mergeData"
  ),
  sexColumn = "Sex",
  outputLogs = file.path(pheno_inputs$tempDir, "logs"),
  outputDir = file.path(pheno_inputs$tempDir, "clockFoundation"),
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = TRUE
)

names(phenoPipelineResult$savedFiles)
names(phenoPipelineResult$savedFiles$timepoints)
basename(unlist(phenoPipelineResult$savedFiles$timepoints[["1"]]))
basename(unlist(phenoPipelineResult$savedFiles[c(
  "combinedPheno",
  "combinedPhenoBeta",
  "betaCSV",
  "phenoCF"
)]))
```

These outputs are the files most commonly consumed by downstream modeling
functions. The combined phenotype-plus-beta object is particularly important
because it becomes the direct input to the GLM and GLMM wrappers.

Arguments used in this example:

- `phenoFile`, `betaPath`, `mPath`, and `cnPath` point to the phenotype table
  and the aligned methylation matrices from the previous workflow stages.
- `SampleID = "Sample_Name"` defines the sample key used during alignment.
- `timeVar = "Timepoint"`, `timepoints = "1,2"`, and
  `combineTimepoints = "1,2"` determine the timepoint-specific outputs and the
  combined longitudinal object.
- `outputPheno`, `outputRData`, `outputRDataMerge`, and `outputDir` define the
  directories used for phenotype exports, metric objects, merged objects, and
  Clock Foundation inputs.
- `saveOutputs = TRUE` is what makes this step useful in a file-based pipeline,
  because later modeling steps consume these written files.

# Write model outputs to disk

The modeling wrappers also separate in-memory results from file outputs. The
next example runs a small GLM analysis and prints the names of the files written
to disk.

```{r glm-pipeline, eval = can_run_glm, message = FALSE, warning = FALSE}
glmPipelineResult <- methylationGLM_T1(
  inputPheno = glm_inputs$inputPath,
  phenotypes = "status",
  covariates = "sex,age",
  factorVars = "status,sex",
  cpgLimit = 2,
  nCores = 1,
  outputLogs = file.path(glm_inputs$tempDir, "logs"),
  outputRData = file.path(glm_inputs$tempDir, "rData", "models"),
  outputPlots = file.path(glm_inputs$tempDir, "figures"),
  significantCpGDir = file.path(glm_inputs$tempDir, "significant"),
  summaryTxtDir = file.path(glm_inputs$tempDir, "summaries"),
  summaryPval = 1,
  significantCpGPval = 1,
  annotationPackage = "IlluminaHumanMethylation450kanno.ilmn12.hg19",
  annotationCols = "Name,chr,pos",
  annotatedGLMOut = file.path(glm_inputs$tempDir, "annotated"),
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = TRUE
)

class(glmPipelineResult$savedFiles)
names(glmPipelineResult$savedFiles)
```

At this stage, the most important saved outputs are the annotated summary table,
the serialized model objects, and the optional text summaries used for
inspection outside R.

Arguments used in this example:

- `inputPheno` points to the merged phenotype-plus-beta object generated by
  `preprocessingPheno()`.
- `phenotypes` lists the outcome variables that will be modeled one at a time.
- `covariates` lists the adjustment variables included in every model.
- `factorVars` identifies the categorical variables that should be treated as
  factors before model fitting.
- `cpgLimit = 2` keeps the example fast by fitting only two CpG models.
- `nCores = 1` keeps the example deterministic and lightweight; production HPC
  runs typically use higher values.
- `annotationPackage` and `annotationCols` control which array annotation is
  merged into the final summary tables.
- `outputRData`, `outputPlots`, `significantCpGDir`, `summaryTxtDir`, and
  `annotatedGLMOut` define where model objects, plots, summaries, and annotated
  results are written.
- `saveOutputs = TRUE` enables the file-based pipeline behaviour expected by
  Makefile and HPC use.

## Model structure and PRS terms

For each CpG, `methylationGLM_T1()` fits a regression model of the form:

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\sum_{k = 1}^{K} \beta_{k + 1} \cdot \text{Covariate}_k +
\varepsilon
$$

where $\text{Methylation}_{CpG_i}$ is the methylation value at CpG $i$,
$\beta_0$ is the intercept, the phenotype term is the variable of interest, and
the remaining fixed effects represent the listed covariates.

If no polygenic risk score is included, a model can be read schematically as:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} =\;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot \text{Age} +
\beta_3 \cdot \text{Sex} +
\beta_4 \cdot \text{Ethnicity} \\
&+
\varepsilon
\end{aligned}
$$

When `prsMap` is supplied, a phenotype-specific PRS term is appended only for
the matching phenotype. For example, the mapping
`"Pheno1:PRS_1,Pheno2:PRS_2"` means that:

- models for `Pheno1` include `PRS_1`
- models for `Pheno2` include `PRS_2`
- phenotypes without a mapping are fit without a PRS term

For a phenotype mapped to a PRS, the model becomes:

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\sum_{k = 1}^{K} \beta_{k + 1} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS} +
\varepsilon
$$

If `interactionTerm` is supplied, the same framework is extended by adding the
requested interaction to the fixed-effect portion of the model.

## Longitudinal mixed-effects structure

The longitudinal wrapper `methylationGLMM_T1T2()` follows the same file-based
pattern, but uses a mixed-effects model with a participant-level random
intercept. In schematic form:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} =\;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot \text{Timepoint} +
\beta_3 \cdot (\text{Phenotype} \times \text{Timepoint}) \\
&+
\sum_{k = 1}^{K} \beta_{k + 3} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS} +
b_{\text{person}} +
\varepsilon
\end{aligned}
$$

where $b_{\text{person}}$ is the subject-specific random intercept. As in the
GLM workflow, the PRS term is included only when the current phenotype is
matched in `prsMap`.

Common arguments in this longitudinal stage are:

- `inputPheno` for the combined longitudinal phenotype-plus-beta object
- `personVar` for the participant identifier used in the random effect
- `timeVar` for the repeated-measures variable
- `phenotypes`, `covariates`, and `factorVars` for the fixed-effect structure
- `prsMap` for phenotype-specific PRS terms
- `cpgLimit` and `nCores` for computational control
- `annotationPackage` and `annotationCols` for annotated summaries

# Typical Makefile use

In a project directory, the Makefile route is usually simpler than calling each
wrapper by hand. Typical commands look like this:

```{r, eval = FALSE}
make f1 MODEL=model1
make f2 MODEL=model1
make f3 MODEL=model1
make f3lme MODEL=model1
make all MODEL=model1
```

The exact target names and variables depend on the Makefile settings extracted
for the project. In practice, `saveOutputs = TRUE` is what makes the individual
steps compatible with this route because later steps consume the files written by
earlier ones.

# Summary

The file-based route is the correct choice when you need:

- a reproducible folder structure,
- files that can be reused by later workflow stages,
- Makefile execution, or
- HPC submission.

The interactive route and the file-based route are therefore complementary:
local use is best for understanding the returned objects, while pipeline use is
best for reproducible multi-step execution.

# Basics

Date the vignette was generated.

```{r, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r, echo = FALSE}
sessionInfo()
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and in
which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve, so it is critical to learn where to ask for help. We would like
to highlight the [Bioconductor support site](https://support.bioconductor.org/)
as the main resource for getting help. Please remember to use the `dnaEPICO`
tag and check the
[older posts](https://support.bioconductor.org/tag/dnaEPICO/). If you want to
receive help, please provide a small reproducible example and your session
information so the source of the problem can be tracked efficiently.
