---
title: "Pipeline Use of dnaEPICO"
author:
  - name: Paul Ruiz
    affiliation:
      - Queensland University of Technology
    email: ruizpint@qut.edu.au
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
  %\VignetteIndexEntry{3. Pipeline Use of dnaEPICO}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)

library(dnaEPICO)

can_run_preprocessing <- requireNamespace("minfiData", quietly = TRUE) &&
    requireNamespace("IlluminaHumanMethylation450kmanifest", quietly = TRUE) &&
    requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)

can_run_sva <- can_run_preprocessing
can_run_glm <- requireNamespace(
    "IlluminaHumanMethylation450kanno.ilmn12.hg19",
    quietly = TRUE
)
can_run_lme <- can_run_glm && requireNamespace("lmerTest", quietly = TRUE)
```

# Introduction

This vignette describes the file-based route in `dnaEPICO`. In this mode, the
same main functions still return structured result objects, but they also
write files to disk when `saveOutputs = TRUE`. This route is appropriate when
you want a reproducible folder structure, Makefile-driven execution, or a
pipeline that can be resumed on a local workstation or an HPC system.

# Citation

The `r Biocpkg("dnaEPICO")` package uses methods from several external
packages. Because no single manuscript describes all components, the guidance
below explains how to cite `dnaEPICO` depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of
[minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd),
[ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and
[wateRmelon](https://github.com/schalkwyk/wateRmelon/blob/master/vignettes/references.bib).

- If you use `make f3 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
    @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools.
- If you use `make f4 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
    @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2.
- If you use `make f3lme MODEL=model1`, please cite @minfi; @minfiEPIC;
    @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @lme4.
- If you use `make all MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix;
    @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2; @lme4.

# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
To read this vignette, we assume that you are already familiar with the general
DNA methylation workflow. If not, we recommend first reading
[this tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html),
which provides a practical introduction to the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor
tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Downstream statistical modelling uses CpG-wise
generalised linear models and longitudinal mixed-effects models fitted with
`lmerTest`/`lme4` or `nlme`. The `nlme` engine can include AR1 or CAR1
within-participant residual correlation. Users are expected to have basic
familiarity with R, command-line execution, and Illumina IDAT file structures.

If you are asking yourself the question "Where do I start using
Bioconductor?", you might be interested in
[this blog post](https://www.bioconductor.org/install/).

## Probe-exclusion reference files

The lists of excluded probes depend on the Illumina methylation-array platform.

- For **Illumina HumanMethylationEPIC v2.0**, use the cross-reactive probe-exclusion file from [Peters et al. (2024)](https://springernature.figshare.com/articles/dataset/Additional_file_8_of_Characterisation_and_reproducibility_of_the_HumanMethylationEPIC_v2_0_BeadChip_for_DNA_methylation_profiling/26689706?file=48531323).

- For **Illumina MethylationEPIC**, also known as the **850k array**, use the probe-exclusion resources from [Pidsley et al. (2016)](https://springernature.figshare.com/articles/dataset/Additional_file_1_Table_S1_of_Critical_evaluation_of_the_Illumina_MethylationEPIC_BeadChip_microarray_for_whole-genome_DNA_methylation_profiling/4405946?file=7132334). The supplementary files commonly used together are [13059_2016_1066_MOESM1_ESM.csv](https://static-content.springer.com/esm/art%3A10.1186%2Fs13059-016-1066-1/MediaObjects/13059_2016_1066_MOESM1_ESM.csv), [13059_2016_1066_MOESM4_ESM.csv](https://static-content.springer.com/esm/art%3A10.1186%2Fs13059-016-1066-1/MediaObjects/13059_2016_1066_MOESM4_ESM.csv), [13059_2016_1066_MOESM5_ESM.csv](https://static-content.springer.com/esm/art%3A10.1186%2Fs13059-016-1066-1/MediaObjects/13059_2016_1066_MOESM5_ESM.csv), and [13059_2016_1066_MOESM6_ESM.csv](https://static-content.springer.com/esm/art%3A10.1186%2Fs13059-016-1066-1/MediaObjects/13059_2016_1066_MOESM6_ESM.csv).

- For **Illumina HumanMethylation450k**, use the cross-reactive and polymorphic probe resources from [Chen et al. (2013)](https://figshare.com/articles/dataset/methylAnalysisDataV3_tar_gz/4800970?file=7896205).

- Multiple probe-exclusion files can be supplied as a semicolon-separated value in `probeExclusionPath`. `dnaEPICO` reads probe IDs from each file, uses `probeExclusionIdColumn` when supplied, or otherwise auto-detects common probe-ID columns such as `ProbeID`, `TargetID`, `IlmnID`, and `Name`. The unique union of all probe IDs is then used to filter the normalised object.

- For **EPICv2**, setting `useEpicV2Manifest = TRUE` also retrieves the expanded Peters et al. manifest from AnnotationHub resource [AH116484](https://bioconductor.org/packages/release/data/annotation/vignettes/EPICv2manifest/inst/doc/EPICv2manifest.html). Probes flagged in selected manifest columns are added to the same exclusion set. By default, probes flagged by `CH_WGBS_evidence`, `CH_BLAT`, or `MissingPos` are removed, while `MismatchPos` is retained unless explicitly enabled.

# Installation

The `r Biocpkg("dnaEPICO")` package depends on `r Biocpkg("minfi")`,
`r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`, among others. Install
dependencies with BiocManager:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("dnaEPICO")

BiocManager::valid()
```

```{r, eval = FALSE, message = FALSE}
library("dnaEPICO")
```

# Extract the example Makefile

The package includes a template Makefile that can be copied into a project
directory and adapted to local paths and model definitions.

```{r makefile-example}
makefilePath <- extractMake(destDir = tempdir(), overwrite = TRUE)
basename(makefilePath)
```

After extraction, the file should be edited so that the project-specific paths,
model labels, and cluster settings match the target analysis environment.

Arguments:

- `destDir = tempdir()` writes the template to a temporary directory so the
    example does not modify the working tree.
- `overwrite = TRUE` allows the template to be replaced if it already exists in
    that temporary location.

# Write preprocessingMinfiEwasWater outputs to disk

The file-based route usually starts with `preprocessingMinfiEwasWater()`. This
step writes the filtered `RGSet`, normalized metric matrices, quality-control
figures, and the `phenoLC.csv` file that will be consumed by `svaEnmix()` and
`preprocessingPheno()`.

### Create example input files

The next chunk shows how to recreate the temporary files used in this example.
It writes a small phenotype table to a temporary directory, copies a small set
of IDAT files, and records the probe-exclusion reference used during probe
filtering.

```{r preprocessing-pipeline-inputs, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
preprocessing_inputs <- dnaEPICO:::exampleMinfiIdatInputsDnaEpico(n = 6L)

names(preprocessing_inputs)
preprocessing_inputs$tempDir
basename(preprocessing_inputs$phenoFile)
head(basename(list.files(preprocessing_inputs$idatFolder, full.names = TRUE)), 4)
basename(preprocessing_inputs$probeExclusionPath)
```

`preprocessing_inputs` is a small list returned by the internal example helper.
`names(preprocessing_inputs)` shows its main elements: `tempDir` is the
temporary working directory, `idatFolder` contains the copied example IDAT
files, `phenoFile` is the phenotype table used by the function, `targets` is the
same phenotype information already loaded into memory, `arrayType` and
`annotationVersion` describe the array platform, and `probeExclusionPath`
points to the probe-exclusion reference used during probe filtering.

```{r preprocessing-pipeline, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
preprocessPipelineResult <- preprocessingMinfiEwasWater(
    phenoFile = preprocessing_inputs$phenoFile,
    idatFolder = preprocessing_inputs$idatFolder,
    outputLogs = file.path(preprocessing_inputs$tempDir, "logs"),
    nSamples = 6,
    SampleID = "Sample_Name",
    arrayType = preprocessing_inputs$arrayType,
    annotationVersion = preprocessing_inputs$annotationVersion,
    scriptLabel = "preprocessingMinfiEwasWater",
    baseDataFolder = file.path(preprocessing_inputs$tempDir, "rData"),
    figureBaseDir = file.path(preprocessing_inputs$tempDir, "figures"),
    detPThreshold = 1,
    normMethods = "quantile",
    sexColumn = "Sex",
    removeSexMismatch = FALSE,
    pvalThreshold = 1,
    chrToRemove = "",
    snpsToRemove = "SBE",
    mafThreshold = 1,
    probeExclusionPath = preprocessing_inputs$probeExclusionPath,
    plotGroupVar = "Sex",
    lcRef = "saliva",
    phenoOrder = "Sample_Name;Sex;Basename;Sentrix_ID;Sentrix_Position",
    lcPhenoDir = file.path(
        preprocessing_inputs$tempDir,
        "data",
        "preprocessingMinfiEwasWater"
    ),
    display = FALSE,
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)

preprocess_paths <- c(
    phenoLC = file.path(
        preprocessing_inputs$tempDir,
        "data",
        "preprocessingMinfiEwasWater",
        "phenoLC.csv"
    ),
    rgset = file.path(
        preprocessing_inputs$tempDir,
        "rData",
        "preprocessingMinfiEwasWater",
        "objects",
        "RGSet.RData"
    ),
    beta = file.path(
        preprocessing_inputs$tempDir,
        "rData",
        "preprocessingMinfiEwasWater",
        "metrics",
        "beta_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData"
    ),
    qc = file.path(
        preprocessing_inputs$tempDir,
        "figures",
        "preprocessingMinfiEwasWater",
        "qc",
        "quality_control(MSet).tiff"
    )
)

class(preprocessPipelineResult)
basename(preprocess_paths)
file.exists(preprocess_paths)
```

The returned object still has class `dnaEPICO_preprocessingMinfiEwasWater`, but
the printed vectors in this chunk focus on the written outputs. In
`preprocess_paths`, `phenoLC` is the phenotype table with estimated cell
proportions, `rgset` is the filtered `RGSet`, `beta` is the saved beta matrix
used later by phenotype preparation, and `qc` is an example quality-control
figure. `basename(preprocess_paths)` shows the file names, while
`file.exists(preprocess_paths)` confirms that each one was written successfully.

Arguments:

- `phenoFile` and `idatFolder` define the phenotype table and the IDAT folder
    used to start the pipeline.
- `SampleID = "Sample_Name"` is the key used to align phenotype rows with the
    methylation objects.
- `arrayType` and `annotationVersion` define the array manifest and annotation
    used by `minfi`.
- `baseDataFolder`, `figureBaseDir`, and `lcPhenoDir` define where serialized
    objects, figures, and `phenoLC.csv` are written.
- `detPThreshold`, `normMethods`, `pvalThreshold`, `chrToRemove`,
    `snpsToRemove`, `mafThreshold`, and `probeExclusionPath` define the main
    preprocessing and filtering choices.
- `sexColumn` identifies reported sex for comparison with methylation-predicted
    sex. `removeSexMismatch = FALSE` retains and reports confirmed mismatches;
    `TRUE` removes samples only when both values are known and disagree. Missing
    or unknown sex values are retained.
- `plotGroupVar`, `lcRef`, and `phenoOrder` control QC grouping, cell-type
    estimation, and the structure of the exported phenotype table.
- `saveOutputs = TRUE` is what turns this function into the first file-producing
    step of the pipeline.

# Write SVA outputs to disk

The next example shows `svaEnmix()` in file-writing mode. The function still
returns a structured result object, but the `savedFiles` element now records the
main output files written to disk.

### Reuse the temporary files written in Step 1

In the file-based route, `svaEnmix()` consumes the `phenoLC.csv` and
`RGSet.RData` files written by `preprocessingMinfiEwasWater()`. The next chunk
shows how those temporary file paths are reconstructed locally.

```{r sva-pipeline-inputs, eval = can_run_sva, message = FALSE, warning = FALSE}
sva_targets_file <- file.path(
    preprocessing_inputs$tempDir,
    "data",
    "preprocessingMinfiEwasWater",
    "phenoLC.csv"
)
sva_rgset_file <- file.path(
    preprocessing_inputs$tempDir,
    "rData",
    "preprocessingMinfiEwasWater",
    "objects",
    "RGSet.RData"
)

basename(c(sva_targets_file, sva_rgset_file))
file.exists(c(sva_targets_file, sva_rgset_file))
```

`sva_targets_file` and `sva_rgset_file` point to the Step 1 outputs:
`phenoLC.csv`, containing the phenotype table and cell-composition estimates,
and `RGSet.RData`, containing the filtered methylation object.

```{r sva-pipeline, eval = can_run_sva, message = FALSE, warning = FALSE}
svaPipelineResult <- svaEnmix(
    phenoFile = sva_targets_file,
    rgsetData = sva_rgset_file,
    outputLogs = file.path(preprocessing_inputs$tempDir, "logs"),
    SampleID = "Sample_Name",
    arrayType = "IlluminaHumanMethylation450k",
    annotationVersion = "ilmn12.hg19",
    SentrixIDColumn = "Sentrix_ID",
    SentrixPositionColumn = "Sentrix_Position",
    ctrlSvaPercVar = 0.90,
    ctrlSvaFlag = 1,
    scriptLabel = "svaEnmix",
    dataBaseDir = file.path(preprocessing_inputs$tempDir, "data"),
    rBaseDir = file.path(preprocessing_inputs$tempDir, "rData"),
    figureBaseDir = file.path(preprocessing_inputs$tempDir, "figures"),
    display = FALSE,
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)

class(svaPipelineResult$savedFiles)
names(svaPipelineResult$savedFiles)
basename(unlist(svaPipelineResult$savedFiles, use.names = FALSE))
```

The returned `savedFiles` object records the main paths written by this step.
`names(svaPipelineResult$savedFiles)` identifies the output groups: `svaRData`
is the serialized surrogate-variable matrix, `svaCSV` is the CSV version of
that matrix, `phenoWithSva` is the same `phenoLC.csv` after the SVA columns are
appended, and `dataDir` / `rDir` are the parent directories used for the saved
outputs. The `basename(...)` call shows the file names generated from those
paths. The phenotype is validated, written to a temporary file, read back, and
then replaced with rollback protection. Sample identifiers, row order,
cell-composition columns, and all other existing columns must remain unchanged.
Existing PC names are rejected before replacement.

Arguments:

- `phenoFile` and `rgsetData` identify the phenotype table and saved `RGSet`
    produced by the earlier preprocessing step.
- `SampleID`, `SentrixIDColumn`, and `SentrixPositionColumn` specify how
    samples and array positions are identified in the phenotype table.
- `ctrlSvaPercVar = 0.90` keeps enough control-derived surrogate variables to
    explain 90% of the control-probe variance.
- `ctrlSvaFlag = 1` enables the ENmix control-based SVA workflow.
- `dataBaseDir` and `rBaseDir` define where file outputs are written when
    `saveOutputs = TRUE`.
- `display = FALSE` and `verbose = FALSE` keep the console output quiet, while
    `logs = TRUE` writes the log files used by the report.

# Write preprocessingPheno outputs to disk

`preprocessingPheno()` is the step that usually prepares the largest number of
pipeline-ready files because it writes timepoint-specific tables, combined
longitudinal tables, and the Clock Foundation export inputs.

### Create example input files

The next chunk shows how to recreate the temporary phenotype and matrix files
used in this example. The helper writes a phenotype table plus aligned beta,
M-value, and copy-number objects to a temporary directory.

```{r preprocessingPheno-pipeline-inputs, message = FALSE, warning = FALSE}
pheno_inputs <- dnaEPICO:::examplePreprocessingPhenoStateDnaEpico()

names(pheno_inputs)
pheno_inputs$tempDir
basename(c(
    pheno_inputs$phenoPath,
    pheno_inputs$betaPath,
    pheno_inputs$mPath,
    pheno_inputs$cnPath
))
```

`pheno_inputs` is a list that bundles the temporary files and the aligned
in-memory objects used in this stage. `names(pheno_inputs)` shows that it
contains the temporary directory, the phenotype table, the saved beta, M-value,
and copy-number files, and precomputed helper objects such as `metricsData`,
`timepointData`, `combinedData`, and `clockFoundation`.

```{r preprocessingPheno-pipeline, message = FALSE, warning = FALSE}
phenoPipelineResult <- preprocessingPheno(
    phenoFile = pheno_inputs$phenoPath,
    betaPath = pheno_inputs$betaPath,
    mPath = pheno_inputs$mPath,
    cnPath = pheno_inputs$cnPath,
    SampleID = "Sample_Name",
    timeVar = "Timepoint",
    timepoints = "1,2",
    combineTimepoints = "1,2",
    outputPheno = file.path(pheno_inputs$tempDir, "data", "preprocessingPheno"),
    outputRData = file.path(
        pheno_inputs$tempDir,
        "rData",
        "preprocessingPheno",
        "metrics"
    ),
    outputRDataMerge = file.path(
        pheno_inputs$tempDir,
        "rData",
        "preprocessingPheno",
        "mergeData"
    ),
    sexColumn = "Sex",
    outputLogs = file.path(pheno_inputs$tempDir, "logs"),
    outputDir = file.path(pheno_inputs$tempDir, "clockFoundation"),
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)

names(phenoPipelineResult$savedFiles)
names(phenoPipelineResult$savedFiles$timepoints)
basename(unlist(phenoPipelineResult$savedFiles$timepoints[["1"]]))
basename(unlist(phenoPipelineResult$savedFiles[c(
    "combinedPheno",
    "combinedPhenoMethylation",
    "betaCSV",
    "phenoCF"
)]))
```

These outputs are the files most commonly consumed by downstream modeling
functions. `names(phenoPipelineResult$savedFiles)` shows the high-level output
groups returned by the function. `names(phenoPipelineResult$savedFiles$timepoints)`
lists the available timepoint-specific subsets. The `basename(...)` calls then
show examples of the concrete files written for one timepoint and for the
combined outputs. In this structure, `combinedPhenoMethylation` is the merged
phenotype-plus-methylation object that becomes the direct input to the GLM and
LME functions. It defaults to beta values, but can be written as M-values or
copy-number values when `methylationScale` is changed. The `betaCSV`, `betaZIP`,
and `phenoCF` outputs remain beta-based Clock Foundation exports.

Arguments:

- `phenoFile`, `betaPath`, `mPath`, and `cnPath` point to the phenotype table
    and the aligned methylation matrices from the previous workflow stages.
- `SampleID = "Sample_Name"` defines the sample key used during alignment.
- `timeVar = "Timepoint"`, `timepoints = "1,2"`, and
    `combineTimepoints = "1,2"` determine the timepoint-specific outputs and the
    combined longitudinal object.
- `methylationScale = "beta"` selects the merged modeling scale. Use `"m"` for
    M-values or `"cn"` for copy-number values. Clock Foundation exports continue
    to use beta values.
- `outputPheno`, `outputRData`, `outputRDataMerge`, and `outputDir` define the
    directories used for phenotype exports, metric objects, merged objects, and
    Clock Foundation inputs.
- `saveOutputs = TRUE` is what makes this step useful in a file-based pipeline,
    because later modeling steps consume these written files.

# Write GLM outputs to disk

The modeling function also separates in-memory results from file outputs. The
next example runs a small GLM analysis and prints the names of the files written
to disk.

### Create example input files

The next chunk shows how to recreate the temporary merged phenotype-plus-beta
input file used by the GLM example.

```{r glm-pipeline-inputs, eval = can_run_glm, message = FALSE, warning = FALSE}
glm_inputs <- dnaEPICO:::exampleMethylationGLMStateDnaEpico()

names(glm_inputs)
glm_inputs$tempDir
basename(glm_inputs$inputPath)
file.exists(glm_inputs$inputPath)
```

`glm_inputs` is a list returned by the GLM example helper. `names(glm_inputs)`
shows that it contains the temporary directory, the saved merged
phenotype-plus-beta input file (`inputPath`), and the corresponding in-memory
objects produced during helper construction: `preparedData`, `modelResults`, and
`modelSummaries`.

```{r glm-pipeline, eval = can_run_glm, message = FALSE, warning = FALSE}
glmPipelineResult <- methylationGLM(
    inputPheno = glm_inputs$inputPath,
    phenotypes = "status",
    covariates = "sex,age",
    factorVars = "status,sex",
    scaleVars = "age",
    cpgLimit = 2,
    nCores = 1,
    outputLogs = file.path(glm_inputs$tempDir, "logs"),
    outputRData = file.path(glm_inputs$tempDir, "rData", "models"),
    outputPlots = file.path(glm_inputs$tempDir, "figures"),
    significantCpGDir = file.path(glm_inputs$tempDir, "significant"),
    summaryTxtDir = file.path(glm_inputs$tempDir, "summaries"),
    summaryPval = 1,
    significantCpGPval = 1,
    annotationPackage = "IlluminaHumanMethylation450kanno.ilmn12.hg19",
    annotationCols = "Name,chr,pos",
    annotatedGLMOut = file.path(glm_inputs$tempDir, "annotated"),
    reportAssetsDir = file.path(
        preprocessing_inputs$tempDir,
        "reports", "example", "assets", "results", "glm_results"
    ),
    display = FALSE,
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)

class(glmPipelineResult$savedFiles)
names(glmPipelineResult$savedFiles)
```

The returned `savedFiles` object records the groups of outputs written by the
GLM step. `names(glmPipelineResult$savedFiles)` includes compact phenotype
summary files, significant-CpG exports, summary
text files, and the annotated results workbook. This is the object to inspect when
you want to confirm which modeling outputs were written and how they are
grouped.

Arguments:

- `inputPheno` points to the merged phenotype-plus-beta object generated by
    `preprocessingPheno()`.
- `phenotypes` lists the outcome variables that will be modeled one at a time.
- `covariates` lists the adjustment variables included in every model.
- `factorVars` identifies the categorical variables that should be treated as
    factors before model fitting.
- `scaleVars` identifies numeric fixed-effect variables to centre and divide by
    their sample standard deviations before fitting.
- Numeric CpG columns are passed unchanged to glm2. Native messages, warnings,
    or errors are stored in `<Phenotype>_Model.Message`; model attempts without
    a returned p-value are counted in workbook metadata and are not added to the
    annotated result table.
- `cpgLimit = 2` keeps the example fast by fitting only two CpG models.
- `nCores = 1` keeps the example deterministic and lightweight. In production,
    `nCores` is a maximum; the effective worker count is also limited by the
    workload, detected CPUs, and available memory.
- `annotationPackage` and `annotationCols` control which array annotation is
    merged into the final summary tables.
- `outputRData`, `outputPlots`, `significantCpGDir`, `summaryTxtDir`, and
    `annotatedGLMOut` define where compact phenotype summaries, plots, text
    summaries, and the
    annotated GLM workbook are written.
- `resumeFromSummary = TRUE` reuses a complete phenotype summary only when it
    was generated from the same input file and model configuration. A phenotype
    without a complete summary restarts from its first CpG.
- `reportAssetsDir` writes the compressed report table directly into the report
    project, leaving the annotated workbook directory free of report sidecars.
- `saveOutputs = TRUE` enables the file-based pipeline behaviour expected by
    Makefile and HPC use.

# Write LME outputs to disk

The longitudinal modeling step follows the same pattern as the GLM function, but
the saved outputs now correspond to mixed-effects models, longitudinal summary
tables, and interaction-specific result files.

### Create example input files

The next chunk shows how to recreate the temporary combined longitudinal
phenotype-plus-beta input file used by the LME example.

```{r lme-pipeline-inputs, eval = can_run_lme, message = FALSE, warning = FALSE}
lme_inputs <- dnaEPICO:::exampleMethylationLMEStateDnaEpico()

names(lme_inputs)
lme_inputs$tempDir
basename(lme_inputs$inputPath)
file.exists(lme_inputs$inputPath)
```

`lme_inputs` plays the same role for the longitudinal example. The helper
returns a list with the temporary directory, the saved combined longitudinal
input file (`inputPath`), and the in-memory objects used to build that example:
`preparedData`, `modelResults`, and `modelSummaries`.

```{r lme-pipeline, eval = can_run_lme, message = FALSE, warning = FALSE}
lmePipelineResult <- methylationLME(
    inputPheno = lme_inputs$inputPath,
    outputLogs = file.path(lme_inputs$tempDir, "logs"),
    outputRData = file.path(lme_inputs$tempDir, "rData", "models"),
    outputPlots = file.path(lme_inputs$tempDir, "figures"),
    personVar = "person",
    timeVar = "Timepoint",
    phenotypes = "score",
    covariates = "sex",
    factorVars = "sex",
    scaleVars = "score",
    cpgLimit = 2,
    nCores = 1,
    summaryPval = 1,
    saveSignificantInteractions = TRUE,
    significantInteractionDir = file.path(
        lme_inputs$tempDir,
        "results",
        "cpgs",
        "methylationLME"
    ),
    significantInteractionPval = 1,
    saveTxtSummaries = TRUE,
    summaryTxtDir = file.path(
        lme_inputs$tempDir,
        "results",
        "summary",
        "methylationLME"
    ),
    annotationPackage = "IlluminaHumanMethylation450kanno.ilmn12.hg19",
    annotationCols = "Name,chr,pos",
    annotatedLMEOut = file.path(lme_inputs$tempDir, "annotated"),
    reportAssetsDir = file.path(
        preprocessing_inputs$tempDir,
        "reports", "example", "assets", "results", "lme_results"
    ),
    display = FALSE,
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)

class(lmePipelineResult$savedFiles)
names(lmePipelineResult$savedFiles)
```

The returned `savedFiles` object records the groups of outputs written by the
LME step. `names(lmePipelineResult$savedFiles)` includes compact phenotype
summary files, summary text files, significant
interaction exports, and the annotated longitudinal results table. This is the
object to inspect when you want to confirm which longitudinal modeling outputs
were written and how they are grouped.

Arguments:

- `inputPheno` points to the combined longitudinal phenotype-plus-beta object.
- `personVar` defines the participant identifier used for the random effect.
- `timeVar` defines the repeated-measures time variable used for longitudinal
    summaries and preprocessing checks.
- `phenotypes`, `covariates`, and optional `interactionTerm` define the
    fixed-effect portion of the mixed model.
- `omnibusTest = TRUE` adds one joint fixed-effect F test per CpG. With an
    interaction, all estimable phenotype-by-interaction coefficients are tested
    together; without an interaction, the phenotype main effect is tested.
- `omnibusDdf` selects `"Satterthwaite"` or `"Kenward-Roger"` denominator
    degrees of freedom. Kenward-Roger testing requires `pbkrtest` and is more
    computationally intensive for an EWAS.
- Numeric CpG columns are passed unchanged to the selected lmerTest/lme4 or
    nlme engine. Native conditions are stored in
    `<Phenotype>_Model.Message`; model attempts without a returned coefficient
    or omnibus p-value are counted in workbook metadata and are not added to
    the annotated result table.
- `cpgLimit = 2` keeps the example short by fitting only two CpG models.
- `nCores = 1` keeps the example lightweight. In production, `nCores` is a
    maximum; the effective worker count is also limited by the workload,
    detected CPUs, and available memory.
- `saveSignificantInteractions`, `significantInteractionDir`, and
    `significantInteractionPval` control the export of interaction-specific
    results.
- `summaryTxtDir` defines where plain-text model summaries are written.
- `annotationPackage`, `annotationCols`, and `annotatedLMEOut` define the
    annotation resources and output location for the final longitudinal summary
    workbook.
- `reportAssetsDir` writes the compressed longitudinal report table directly
    into the report project rather than beside the workbook.
- `resumeFromSummary = TRUE` reuses a complete phenotype summary only when it
    was generated from the same input file and model configuration. A phenotype
    without a complete summary restarts from its first CpG.
- `saveOutputs = TRUE` enables the file-based longitudinal workflow expected by
    Makefile and HPC use.

For a full-cohort longitudinal interaction analysis, the following structure
fits a participant-level random intercept and returns one joint interaction
p-value per CpG in addition to the component coefficient p-values:

```{r lme-omnibus-example, eval = FALSE}
omnibusResult <- methylationLME(
    inputPheno = "rData/model/preprocessingPheno/mergeData/phenoBT1T2T3.RData",
    personVar = "Participant",
    timeVar = "Timepoint",
    phenotypes = "Timepoint",
    covariates = "Age,Sex,PC1,PC2,PC3",
    factorVars = "Timepoint,Profession,Sex",
    interactionTerm = "Profession",
    omnibusTest = TRUE,
    omnibusDdf = "Satterthwaite"
)
```

This model uses the fixed effects `Timepoint * Profession` together with
`(1 | Participant)`. A significant omnibus result indicates that at least one
timepoint contrast differs by profession; the component coefficients are used
to identify its direction and location.

## Generate the report from the example outputs

After the file-based examples above have been run, `dnamReport()` can assemble
the phenotype table, ENmix control figures, quality-control figures, batch
effect figures, metrics figures, model annotation tables, and logs into one
website report. The example below uses the concrete output paths created by the
previous chunks.

The preprocessing and SVA examples share `preprocessing_inputs$tempDir`, while
the GLM and LME examples write model outputs into their own temporary
directories. Because `dnamReport()` accepts one path per tab, those outputs can
be passed directly.

When running this section from an R session, `dnamReport()` still needs the
Quarto command line interface because the function renders the `.qmd` files into
the final website. You can check whether R can find Quarto with:

```{r quarto-check-r, eval = FALSE}
Sys.which("quarto")
system2("quarto", "--version")
```

If `Sys.which("quarto")` returns an empty string, install the Quarto command
line interface from the
[Quarto get-started page](https://quarto.org/docs/get-started/) before
rendering the report. The R package `quarto` provides R helper functions for an
existing Quarto installation; it does not replace the Quarto command line
interface used by `dnamReport()`.

If Quarto is installed but not on `PATH`, set the executable path before calling
`dnamReport()`:

```{r quarto-path-r, eval = FALSE}
Sys.setenv(QUARTO_BIN = "/full/path/to/quarto")
```

```{r dnamReport-pipeline-example, eval = FALSE}
report_log_dir <- file.path(preprocessing_inputs$tempDir, "logs", "report")
dir.create(report_log_dir, recursive = TRUE, showWarnings = FALSE)

report_logs <- c(
    file.path(
        preprocessing_inputs$tempDir,
        "logs",
        "log_preprocessingMinfiEwasWater.txt"
    ),
    file.path(pheno_inputs$tempDir, "logs", "log_preprocessingPheno.txt"),
    file.path(preprocessing_inputs$tempDir, "logs", "log_svaEnmix.txt"),
    file.path(glm_inputs$tempDir, "logs", "log_methylationGLM.txt"),
    file.path(lme_inputs$tempDir, "logs", "log_methylationLME.txt")
)

existing_logs <- report_logs[file.exists(report_logs)]
if (length(existing_logs)) {
    file.copy(existing_logs, report_log_dir, overwrite = TRUE)
}

reportResult <- dnamReport(
    outputDir = file.path(preprocessing_inputs$tempDir, "reports", "example"),
    phenoTab = file.path(
        preprocessing_inputs$tempDir,
        "data",
        "preprocessingMinfiEwasWater",
        "phenoLC.csv"
    ),
    enmixTab = file.path(
        preprocessing_inputs$tempDir,
        "figures",
        "preprocessingMinfiEwasWater",
        "enmix"
    ),
    qcTab = file.path(
        preprocessing_inputs$tempDir,
        "figures",
        "preprocessingMinfiEwasWater",
        "qc"
    ),
    svaTab = file.path(preprocessing_inputs$tempDir, "figures", "svaEnmix"),
    metricTab = file.path(
        preprocessing_inputs$tempDir,
        "figures",
        "preprocessingMinfiEwasWater",
        "metrics"
    ),
    glmTab = glmPipelineResult$savedFiles$annotatedGLM,
    lmeTab = lmePipelineResult$savedFiles$annotatedLME,
    logTab = report_log_dir,
    detPPath = file.path(
        preprocessing_inputs$tempDir,
        "rData",
        "preprocessingMinfiEwasWater",
        "qc",
        "detP_RGSet.RData"
    ),
    detPThreshold = 0.01,
    verbose = FALSE,
    logs = TRUE
)

reportResult$outputFile
```

## GLM model structure and PRS terms

For each CpG, `methylationGLM()` fits a regression model of the form:

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\sum_{k = 1}^{K} \beta_{k + 1} \cdot \text{Covariate}_k +
\varepsilon
$$

where $\text{Methylation}_{CpG_i}$ is the methylation value at CpG $i$,
$\beta_0$ is the intercept, the phenotype term is the variable of interest, and
the remaining fixed effects represent the listed covariates.

Setting `omnibusTest = TRUE` adds one joint Wald F test per CpG using
`car::linearHypothesis()`. Without an interaction, the complete phenotype term
is tested. With an interaction, the complete phenotype-by-interaction term is
tested. One-degree-of-freedom terms, including a numeric phenotype entered as
one linear term, are also tested and reproduce the corresponding coefficient
p-value. Omnibus p-values are adjusted across valid CpGs within each phenotype
and tested term using `padjmethod`.

If no polygenic risk score is included, a model can be read schematically as:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} =\;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot \text{Age} +
\beta_3 \cdot \text{Sex} +
\beta_4 \cdot \text{Ethnicity} \\
&+
\varepsilon
\end{aligned}
$$

When `prsMap` is supplied, a phenotype-specific PRS term is appended only for
the matching phenotype. For example, the mapping
`"Pheno1:PRS_1,Pheno2:PRS_2"` means that:

- models for `Pheno1` include `PRS_1`
- models for `Pheno2` include `PRS_2`
- phenotypes without a mapping are fit without a PRS term

For a phenotype mapped to a PRS, the model becomes:

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\sum_{k = 1}^{K} \beta_{k + 1} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS} +
\varepsilon
$$

If `interactionTerm` is supplied, the same framework is extended by adding the
requested interaction to the fixed-effect portion of the model.

## Longitudinal mixed-effects structure

The longitudinal function `methylationLME()` follows the same file-based
pattern, but uses a mixed-effects model with a participant-level random
intercept. In schematic form:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} =\;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot (\text{Phenotype} \times \text{Interaction}) +
\sum_{k = 1}^{K} \beta_{k + 2} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS} +
b_{\text{person}} +
\varepsilon
\end{aligned}
$$

where $b_{\text{person}}$ is the subject-specific random intercept. As in the
GLM workflow, the PRS term is included only when the current phenotype is
matched in `prsMap`.

Common arguments in this longitudinal stage are:

- `inputPheno` for the combined longitudinal phenotype-plus-beta object
- `personVar` for the participant identifier used in the random effect
- `timeVar` for repeated-measures summaries and preprocessing checks
- `phenotypes`, `covariates`, and optional `interactionTerm` for the fixed-effect
    structure
- `prsMap` for phenotype-specific PRS terms
- `cpgLimit` and `nCores` for computational control
- `annotationPackage` and `annotationCols` for annotated summaries

# Makefile use

In a project directory, the Makefile route is simpler than calling each
function by hand. A minimal project layout looks like this:

The workflow requires GNU Make 4.3 or later and `Rscript` on the system path.
Quarto is required when a report target is requested. The exported rules use
relative project paths, including when the project directory contains spaces.

```text
my_project/
    Makefile
    metadata/
    pheno_model1.csv
    data/
    preprocessingMinfiEwasWater/
        idats/
        <IDAT files>
```

The exported template should then be edited so that at least the model name,
phenotype path, and directory variables match the project. A minimal example is:

```make
MODEL = model1
PHENO_FILE = metadata/pheno_model1.csv
LOGS_DIR = logs
DATA_DIR = data
RDATA_DIR = rData
RESULTS_DIR = results
FIGURES_DIR = figures
```

With that structure in place, a typical sequence of commands looks like this:

```bash
make step1 MODEL=model1
make step2 MODEL=model1
make step3 MODEL=model1
make f3 MODEL=model1
make f4 MODEL=model1
make f3lme MODEL=model1
make all MODEL=model1
make status MODEL=model1
make clean MODEL=model1
```

These targets correspond to the grouped outputs defined in the exported rules:

- `make step1 MODEL=model1` runs preprocessing only.
- `make step2 MODEL=model1` runs the SVA step using the files written by Step 1.
- `make step3 MODEL=model1` runs phenotype preparation after Step 2 has safely
    appended PC columns to `phenoLC.csv`.
- `make f3 MODEL=model1` runs preprocessing, SVA, phenotype preparation, and a
    report containing only those sections.
- `make f4 MODEL=model1` runs the same route, adds the GLM step, and generates a
    report containing the preprocessing and GLM sections.
- `make f3lme MODEL=model1` runs the same route, adds the longitudinal LME step,
    and generates a report containing the preprocessing and LME sections.
- `make all MODEL=model1` runs the full pipeline, including both modeling steps and the report.
- `make status MODEL=model1` checks which expected output files are already present.
- `make clean MODEL=model1` removes the files generated for the selected model while keeping the shared raw input area.

If several models are declared in `MODELS`, the grouped multi-model targets can
be used instead:

```bash
make f3_models
make f4_models
make f3lme_models
make models
```

These grouped targets loop over every model listed in `MODELS`:

- `make f3_models` runs the `f3` route for each declared model.
- `make f4_models` runs the `f4` route for each declared model.
- `make f3lme_models` runs the `f3lme` route for each declared model.
- `make models` runs the full `all` pipeline for each declared model.

Two maintenance targets are also useful during routine work:

```bash
make status MODEL=model1
make clean MODEL=model1
make clean_models
```

- `make status MODEL=model1` reports which expected outputs are already present for one model.
- `make clean MODEL=model1` removes the generated outputs for one model.
- `make clean_models` removes the generated outputs for every model listed in
    `MODELS`.

The exact target names and variables depend on the Makefile settings extracted
for the project. In practice, `saveOutputs = TRUE` is what makes the individual
steps compatible with this route because later steps consume the files written
by earlier ones. Step 2 creates
`data/<model>/svaEnmix/.sva_complete` only after the updated `phenoLC.csv` has
been validated and installed. Downstream targets depend on both files, so Make
reruns SVA when Step 1 replaces `phenoLC.csv` and skips SVA when the completed
inputs are unchanged.

## Report generation

The exported Makefile contains a report rule whose output is
`$(STEP6)/$(MODEL)/docs/index.html`. With the default directory settings, this
becomes `reports/<model>/docs/index.html`. Users usually do not need to call
that file target directly: the grouped targets above depend on it, so Make
refreshes the report automatically after the required pipeline outputs are
available.

For example, the `f3` route builds the report from the Data, ENmix QC, Quality
Control, Batch Effect, Metrics, Report, and Logs inputs. The `f4` route adds only
the GLM page and GLM log, the `f3lme` route adds only the LME page and LME log,
and `all` includes both modeling sections. Each report contains the model
sections defined by its route.

The report target reads the same types of outputs generated in the examples
above, organised under the Makefile project layout: the phenotype table in
`data/`, quality-control and metric figures in `figures/`, the detection
P-value object in `rData/`, model annotation tables in `data/`, and workflow
logs in `logs/`.

The generated site is opened from the report target path:

```text
$(STEP6)/$(MODEL)/docs/index.html
```

Internally, the Makefile calls `dnamReport()` with one argument per tab, using
the project paths defined by the Makefile variables. The report target requires
the Quarto command line interface in the same environment where `make` is run.
If Quarto is not already available, install the Quarto command line interface
from the [Quarto get-started page](https://quarto.org/docs/get-started/), or
use the installation method provided by the local computing environment. Before
launching the report target, check:

```bash
quarto --version
```

If Quarto is installed outside `PATH`, export its executable path before running
`make`:

```bash
export QUARTO_BIN=/full/path/to/quarto
```

# Full exported Makefile

The full Makefile exported by `extractMake()` is shown below. This is the
template that should be adapted for a project before running the pipeline in the
target analysis environment.

```{r makefile-template-full, echo = FALSE, results = "asis"}
makefile_lines <- readLines(makefilePath, warn = FALSE)
cat("```make\n", paste(makefile_lines, collapse = "\n"), "\n```\n", sep = "")
```

# Makefile section guide

The exported Makefile is organised into sections so that project-specific
choices can be edited without changing the pipeline rules themselves.

## User configuration

- `MAKEFLAGS += --output-sync` keeps parallel Make output grouped by recipe,
    which makes logs easier to read.

## Models selection and parallel run

- `MODEL` defines a single pipeline configuration to run.
- `MODELS` defines a set of configurations that can be launched in parallel by
    the grouped Make targets.

In single-model use, commands such as `make f4 MODEL=model1` run one workflow.
In multi-model use, commands such as `make f4_models` evaluate the same template
for each name listed in `MODELS`.

## Per-model overrides

- `PHENO_FILE` is reassigned according to the selected `MODEL`.

This block is useful when different cohorts, case definitions, or phenotype
encodings should reuse the same analysis pipeline without duplicating the full
Makefile.

## Directories

- `LOGS_DIR`, `DATA_DIR`, `RDATA_DIR`, `RESULTS_DIR`, and `FIGURES_DIR` define
    the top-level folder structure.
- `STEP1` to `STEP6` define the canonical names of the major pipeline stages.
- `METRICS_DIR`, `IDAT_DIR`, `OBJ_DIR`, `MERGE_DIR`, `MODEL_DIR`, `CPG_DIR`,
    `SUMMARY_DIR`, `ENMIX_DIR`, `SVA_DIR`, and `QC_DIR` define the subdirectory
    names used by the rules file.
- `R_DIR` optionally points to a custom R library path on shared systems. Its
    default value is `NULL`, which uses the default library path.

These variables control where files are written and how paths are composed
across all stages of the pipeline.

## Global parameters

These parameters are shared across multiple workflow steps.

### Step 1 to Step 5 shared parameters

- `PHENO_FILE`: input phenotype table used to start the workflow.
- `SEP_TYPE`: optional separator used when reading text files; `NULL` uses the
    default comma separator.
- `SAMPLE_ID`: phenotype column used to align samples.
- `N_SAMPLES`: optional sample subset for testing; `NA` uses all samples.
- `ARRAY_TYPE` and `ANNOTATION_VERSION`: array platform and annotation build.
- `IDAT_FORCE`: whether `minfi::read.metharray.exp()` should force reading
    selected IDAT files with different internal array sizes; the default is
    `FALSE`.
- `TIFF_WIDTH`, `TIFF_HEIGHT`, and `TIFF_RES`: default plot dimensions and
    resolution.
- `SEX_COLUMN`: phenotype column used for sex-based checks and prediction.
- `SENTRIX_ID_COLUMN` and `SENTRIX_POSITION_COLUMN`: chip-position metadata
    used by the SVA stage.
- `BASENAME_COLUMN`: column holding the array basename.
- `TIME_VAR`: phenotype column used for repeated-measures or longitudinal
    workflows.
- `METHYLATION_SCALE`: methylation metric used in merged modeling tables;
    choose `beta`, `m`, or `cn`. The default is `beta`; Clock Foundation exports
    continue to use beta values.
- `PHENO_ORDER`: desired order of key phenotype columns in exported tables.

### Step 4 and Step 5 modeling parameters

- `COVARIATES_GLM` and `COVARIATES_LME`: comma-separated adjustment variables
    included in the GLM and LME analyses, respectively.
- `FACTOR_VARS_GLM` and `FACTOR_VARS_LME`: variables treated as categorical in
    the GLM and LME analyses, respectively.
- `SCALE_VARS_GLM` and `SCALE_VARS_LME`: numeric fixed-effect variables centred
    and divided by their sample standard deviations in the corresponding
    analysis; `NULL` keeps the variables on their original scales.
- `CPG_PREFIX`: prefix used to identify CpG columns.
- `CPG_LIMIT`: optional cap on the number of CpGs processed; `NA` uses all.
- `PRS_MAP`: optional mapping from phenotype names to PRS variables; `NULL`
    omits PRS terms from the models.
- `SUMMARY_PVAL`: optional p-value cutoff for summary tables; `NA` keeps all
    summary rows.
- `N_CORES`: number of cores used within each R job.
- `SAVE_TXT_SUMMARIES`: whether plain-text summaries should be written.
- `RESUME_FROM_SUMMARY`: whether complete phenotype summaries generated from
    the same input files and model configurations may be reused after a model
    stage stops.
- `CHUNK_SIZE`: number of CpGs processed per chunk; `NULL` lets the R helpers
    choose the chunk size automatically. Chunks bound in-memory computation;
    they are not saved as restart checkpoints.
- `FDR_THRESHOLD`: false-discovery-rate threshold used for filtering.
- `PADJ_METHOD`: multiple-testing correction method.
- `ANNOTATION_PACKAGE`: array annotation package used for annotated summaries.
- `ANNOTATION_COLS`: annotation fields to merge into output tables.

## Step 1 parameters

- `QC_CUTOFF`: sample-quality threshold applied during preprocessing.
- `DET_PTYPE`: detection p-value method.
- `DET_PTHRESHOLD`: detection p-value cutoff.
- `PVAL_THRESHOLD`: probe-level p-value filter applied after preprocessing.
- `CHR_TO_REMOVE`: chromosomes to exclude from downstream matrices.
- `SNPS_TO_REMOVE`: SNP-related probe categories to remove.
- `PROBE_EXCLUSION_FILE`: semicolon-separated probe-exclusion reference files.
- `PROBE_EXCLUSION_ID_COLUMN`: optional column or semicolon-separated columns
    containing probe IDs; `NULL` auto-detects each reference file.
- `USE_EPICV2_MANIFEST`: whether to also remove EPICv2 probes flagged in the
    Peters et al. expanded manifest from AnnotationHub resource `AH116484`.
- `EPICV2_MANIFEST_CH_WGBS_EVIDENCE`, `EPICV2_MANIFEST_CH_BLAT`,
    `EPICV2_MANIFEST_MISSING_POS`, and `EPICV2_MANIFEST_MISMATCH_POS`: manifest
    flags included in the EPICv2 probe-exclusion set.
- `MAF_THRESHOLD`: minor-allele-frequency threshold used during SNP filtering.
- `PLOT_GROUP_VAR`: phenotype variable used for grouped QC plots.
- `LC_REF`: reference panel used for cell-type estimation.
- `REMOVE_SEX_MISMATCH`: `FALSE` retains and reports confirmed sex mismatches;
    `TRUE` removes samples only when reported and methylation-predicted sex are
    both known and disagree. Samples with missing or unknown sex are retained.

These values control the main preprocessing and filtering choices in
`preprocessingMinfiEwasWater()`.

## Step 2 parameters

- `CTRL_SVA_PERC_VAR`: target proportion of control-probe variance explained by
    the retained surrogate variables.
- `CTRL_SVA_FLAG`: enables the control-based ENmix SVA workflow.

These values determine how `svaEnmix()` estimates and retains surrogate
variables. The completed step updates the Step 1 `phenoLC.csv` in place and
records successful completion with `.sva_complete`; it does not create a second
phenotype file.

## Step 3 parameters

- `TIMEPOINTS`: comma-separated timepoints exported separately.
- `COMBINE_TIMEPOINTS`: comma-separated timepoints combined into the
    longitudinal object.
- `METHYLATION_SCALE`: methylation metric used for the merged GLM/LME input.
    `Beta` writes `phenoB*`, `M` writes `phenoM*`, and `CN` writes `phenoCN*`;
    scale input is case-insensitive.

These values determine how `preprocessingPheno()` splits and recombines the
phenotype-methylation data.

## Step 4 parameters

- `PHENOTYPES_GLM`: comma-separated phenotype variables that will be modelled
    one at a time by `methylationGLM()`.
- `GLM_LIBS`: GLM implementation used by `methylationGLM()`.
- `INTERACTION_GLM`: optional interaction term added to the fixed-effect part
    of the GLM; `NULL` omits the interaction.
- `GLM_OMNIBUS_TEST`: whether one joint Wald F test is reported for each CpG's
    phenotype or phenotype-by-interaction term.
- `SUMMARY_RESIDUAL_SD`: whether residual standard deviations are reported in
    summaries.
- `SAVE_SIGNIFICANT_CPGS`: whether tables of significant CpGs are exported.
- `SIGNIFICANT_CPG_PVAL`: p-value threshold used for those significant-CpG
    exports.
- Annotated GLM rows require at least one returned coefficient or omnibus
    p-value. Native glm2 and omnibus-test conditions are written to the
    phenotype-specific `Model.Message` column and aggregate availability counts
    are written to metadata.

## Step 5 parameters

- `PHENOTYPES_LME`: comma-separated phenotype variables that will be modelled
    one at a time by `methylationLME()`.
- `PERSON_VAR`: participant identifier used as the random effect grouping
    factor.
- `LME_LIBS`: mixed-effects libraries used by `methylationLME()`.
- `LME_CORRELATION_STRUCTURE`: residual correlation structure for the `nlme`
    LME backend; use `none`, `AR1`, or `CAR1`.
- `LME_CORRELATION_VAR`: variable used to order repeated observations within
    participant for `AR1` or `CAR1`; required for those structures.
- `INTERACTION_LME`: optional interaction term used in the longitudinal model;
    `NULL` omits the interaction.
- `LME_OMNIBUS_TEST`: whether lmerTest/lme4 models calculate one joint test for
    the selected phenotype main effect or phenotype-by-interaction term.
- `LME_OMNIBUS_DDF`: denominator degrees-of-freedom method for the joint test;
    use `Satterthwaite` or `Kenward-Roger`.
- `SAVE_SIGNIFICANT_INTERACTIONS`: whether significant interaction tables are
    exported.
- `SIGNIFICANT_INTERACTION_PVAL`: p-value threshold for those exported
    interaction results.
- Annotated LME rows require at least one returned coefficient or omnibus
    p-value. Native lmerTest/lme4 or nlme conditions are written to the
    phenotype-specific `Model.Message` column and aggregate availability counts
    are written to metadata.

## Argument normalisation

- The GLM and LME covariate, factor, and scaling variables are converted
    directly with `optional_text_arg`. `PRS_MAP_ARG`, `SEP_TYPE_ARG`,
    `R_DIR_ARG`, `PROBE_EXCLUSION_ID_COLUMN_ARG`, `INTERACTION_GLM_ARG`, and
    `INTERACTION_LME_ARG` provide the same conversion for other Makefile
    settings.

This section is especially important for values such as `NULL`, because it
prevents the literal string `"NULL"` from being interpreted as a real model
value inside the main functions.

The template convention is therefore: use `NA` for optional numeric limits or
filters where the meaning is "use all" or "keep all", and use `NULL` when an
optional separator, variable, path, mapping, or model term should be absent.
`CHUNK_SIZE` uses `NULL` for automatic chunking because it disables a
user-specified tuning value rather than applying a filter.

## Include pipeline from dnaEPICO

- `DNAPIPE_MK` resolves the packaged `Makefile.rules.pipeline` file.
- `include $(DNAPIPE_MK)` imports the actual pipeline rules after the user
    configuration has been defined.

This separation is what allows the template Makefile to stay compact while the
package keeps the rule implementation in one maintained location.

# Summary

The file-based route is the correct choice when you need:

- a reproducible folder structure,
- files that can be reused by later workflow stages,
- Makefile execution, or
- repeated analyses across one or more models.

The interactive route and the file-based route are therefore complementary:
local use is best for understanding the returned objects, while pipeline use is
best for reproducible multi-step execution.

# Basics

Date the vignette was generated.

```{r, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r, echo = FALSE}
sessionInfo()
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and in
which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve, so it is critical to learn where to ask for help. We would like
to highlight the [Bioconductor support site](https://support.bioconductor.org/)
as the main resource for getting help. Please remember to use the `dnaEPICO`
tag and check the
[older posts](https://support.bioconductor.org/tag/dnaEPICO/). If you want to
receive help, please provide a small reproducible example and your session
information so the source of the problem can be tracked efficiently.

# References
