---
title: "Pipeline Overview"
author:
  - name: Paul Ruiz
    affiliation:
      - Queensland University of Technology
    email: ruizpint@qut.edu.au
  - name: Divya Mehta
    affiliation:
      - Queensland University of Technology
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
vignette: >
  %\VignetteIndexEntry{1. Pipeline Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL
)

library(dnaEPICO)
```

# Introduction

This vignette provides a visual overview of the main `dnaEPICO`
pipelines roles. It complements the local-use and pipeline vignettes
by showing how the major functions organise their inputs, internal
processing stages, and outputs.

The diagrams are intended as orientation maps. They are useful when deciding
which function to run, checking how outputs move between stages, or explaining
the pipeline structure to collaborators. For detailed arguments and executable
examples, use the local-use and pipeline-use vignettes.

Read each overview as follows:

- The arrows show the expected flow from inputs to derived objects and files.
- The grouped regions show the main analysis tasks performed inside a function.
- The final nodes show the outputs that are returned to R or written to disk.
- The same function can be used interactively or as part of a file-based
  pipeline, depending on the value of `saveOutputs`.

# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
To read this vignette, we assume that you are already familiar with the general
DNA methylation pipeline. If not, we recommend first reading
[this tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html),
which provides a practical introduction to the main concepts and analysis
steps.

Preprocessing and quality control are performed using established
Bioconductor tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Downstream statistical modelling relies on base R
and CRAN frameworks, including generalised linear models and linear
mixed-effects models. Users are expected to have basic familiarity with R,
Bioconductor pipelines, command-line execution, and Illumina IDAT file
structures.

If you are asking yourself the question "Where do I start using
Bioconductor?", you might be interested in
[this blog post](https://www.bioconductor.org/install/).

```{r overview-helpers, echo = FALSE, results = "asis"}
cat("
<style>
.overview-figure {
  margin: 1.5rem 0 2rem;
}

.overview-figure-title {
  font-weight: 700;
  margin: 0 0 0.5rem;
}

.overview-controls {
  display: flex;
  flex-wrap: wrap;
  gap: 8px;
  margin: 0 0 0.5rem;
}

.overview-zoom-button {
  background: #ffffff;
  border: 1px solid #2c7fb8;
  border-radius: 6px;
  color: #2c7fb8;
  cursor: pointer;
  font-size: 0.875rem;
  line-height: 1.2;
  padding: 5px 9px;
}

.overview-zoom-button:hover,
.overview-zoom-button:focus {
  background: #edf6fb;
}

.overview-zoom-value {
  align-items: center;
  color: #4b5563;
  display: inline-flex;
  font-size: 0.875rem;
  min-width: 44px;
}

.overview-svg-frame {
  background: #ffffff;
  border: 1px solid #d9e2ec;
  border-radius: 6px;
  max-height: 680px;
  overflow: auto;
  padding: 10px;
}

.overview-svg-frame svg {
  height: auto;
  max-width: none;
  width: calc(1600px * var(--overview-scale, 1));
}

.overview-figure:fullscreen {
  background: #ffffff;
  margin: 0;
  overflow: hidden;
  padding: 16px;
}

.overview-figure:fullscreen .overview-svg-frame {
  height: calc(100vh - 88px);
  max-height: none;
}

@media (max-width: 768px) {
  .overview-svg-frame {
    max-height: 520px;
  }

  .overview-svg-frame svg {
    width: calc(1200px * var(--overview-scale, 1));
  }
}
</style>
<script>
function setOverviewZoom(figureId, scale) {
  var figure = document.getElementById(figureId);
  if (!figure) {
    return;
  }

  var next = Math.max(0.5, Math.min(3, scale));
  figure.dataset.scale = next.toFixed(2);
  figure.style.setProperty('--overview-scale', next.toString());

  var label = figure.querySelector('.overview-zoom-value');
  if (label) {
    label.textContent = Math.round(next * 100) + '%';
  }
}

function changeOverviewZoom(figureId, delta) {
  var figure = document.getElementById(figureId);
  var current = figure ? parseFloat(figure.dataset.scale || '1') : 1;
  setOverviewZoom(figureId, current + delta);
}

function openOverviewFullscreen(figureId) {
  var figure = document.getElementById(figureId);
  if (!figure || !figure.requestFullscreen) {
    return;
  }

  figure.requestFullscreen();
}
</script>
")

overview_svgs <- c(
  preprocessingMinfiEwasWater =
    "preprocessingMinfiEwasWater_pipeline_overview.svg",
  svaEnmix =
    "svaEnmix_pipeline_overview.svg",
  preprocessingPheno =
    "preprocessingPheno_pipeline_overview.svg",
  methylationGLM_T1 =
    "methylationGLM_T1_pipeline_overview.svg",
  methylationGLMM_T1T2 =
    "methylationGLMM_T1T2_pipeline_overview.svg",
  dnamReport =
    "dnamReport_pipeline_overview.svg"
)

html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x
}

svg_to_html <- function(svg_path, title = NULL) {
  if (!file.exists(svg_path)) {
    stop("Missing overview SVG: ", svg_path)
  }

  svg_lines <- readLines(svg_path, warn = FALSE, encoding = "UTF-8")
  svg_start <- grep("<svg\\b", svg_lines)[1]

  if (is.na(svg_start)) {
    stop("No <svg> tag found in: ", svg_path)
  }

  svg_markup <- paste(svg_lines[svg_start:length(svg_lines)], collapse = "\n")
  figure_id <- gsub("[^A-Za-z0-9_-]", "-", tools::file_path_sans_ext(
    basename(svg_path)
  ))

  cat(sprintf(
    '<div class="overview-figure" id="%s" data-scale="1">\n',
    figure_id
  ))
  cat(sprintf(
    '<p class="overview-figure-title">%s</p>\n',
    html_escape(title)
  ))
  cat('<div class="overview-controls">\n')
  cat(sprintf(
    '<button class="overview-zoom-button" type="button" onclick="changeOverviewZoom(\'%s\', 0.25)">Zoom +</button>\n',
    figure_id
  ))
  cat(sprintf(
    '<button class="overview-zoom-button" type="button" onclick="changeOverviewZoom(\'%s\', -0.25)">Zoom -</button>\n',
    figure_id
  ))
  cat(sprintf(
    '<button class="overview-zoom-button" type="button" onclick="setOverviewZoom(\'%s\', 1)">Reset</button>\n',
    figure_id
  ))
  cat(sprintf(
    '<button class="overview-zoom-button" type="button" onclick="openOverviewFullscreen(\'%s\')">Full screen</button>\n',
    figure_id
  ))
  cat('<span class="overview-zoom-value">100%</span>\n')
  cat('</div>\n')
  cat('<div class="overview-svg-frame">\n')
  cat(svg_markup)
  cat('\n</div>\n</div>\n')
}
```

# Pipelines roles

The sections below describe the role of each main function in the package pipeline.
They are ordered according to the usual analysis path: preprocessing,
surrogate-variable estimation, phenotype preparation, modelling, longitudinal
modelling, and report generation.

## preprocessingMinfiEwasWater()

`preprocessingMinfiEwasWater()` reads the phenotype table and IDAT files, builds the methylation
objects, performs quality control and normalisation, filters probes, and
estimates cell composition.

Its role in the pipeline is to create analysis-ready methylation data:

- Main inputs: phenotype metadata, IDAT files, array annotation, and filtering
  settings.
- Main processing tasks: import, quality control, normalisation, probe
  filtering, metric extraction, and cell-composition estimation.
- Main outputs: filtered phenotype data, `RGSet`, beta values, M-values,
  copy-number values, quality-control figures, and `phenoLC`.

```{r preprocessingMinfiEwasWater-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["preprocessingMinfiEwasWater"]]
)
```

## svaEnmix()

`svaEnmix()` estimates surrogate variables from control-probe information and
adds them to the phenotype table. This step helps represent technical
variation that may otherwise influence downstream association models.

Its role is to prepare covariates for batch and technical adjustment:

- Main inputs: the phenotype table from preprocessing and the saved `RGSet`.
- Main processing tasks: control-probe extraction, surrogate-variable
  estimation, association checks with array-position metadata, and phenotype
  merging.
- Main outputs: surrogate-variable matrices, an updated phenotype table, and
  diagnostic summaries.

```{r svaEnmix-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["svaEnmix"]])
```

## preprocessingPheno()

`preprocessingPheno()` aligns phenotype information with methylation metrics.
It prepares timepoint-specific data, combines longitudinal records, and creates export-ready files for external methylation-age tools.

Its role is to organise samples and methylation matrices for modelling:

- Main inputs: phenotype data, beta values, M-values, copy-number values, and
  timepoint settings.
- Main processing tasks: sample alignment, timepoint splitting, longitudinal
  merging, and Clock Foundation export preparation.
- Main outputs: timepoint-specific phenotype-methylation tables, combined
  longitudinal tables, and Clock Foundation input files.

```{r preprocessingPheno-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["preprocessingPheno"]])
```

## methylationGLM_T1()

`methylationGLM_T1()` fits cross-sectional methylation association models. It
is designed for analyses where one phenotype is tested against CpG-level
methylation while adjusting for selected covariates.

Its role is to run single-timepoint association testing:

- Main inputs: a phenotype-methylation table, phenotype variables,
  covariates, optional PRS mappings, and annotation settings.
- Main processing tasks: model preparation, CpG-GLM model fitting, diagnostic
  plotting, result summarisation, multiple-testing adjustment, and annotation.
- Main outputs: model objects, summary tables, diagnostic figures,
  significant-CpG exports, and annotated workbooks.

```{r methylationGLM_T1-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["methylationGLM_T1"]])
```

## methylationGLMM_T1T2()

`methylationGLMM_T1T2()` fits longitudinal mixed-effects models. It supports
repeated-measures designs by including participant-level random effects and
timepoint-related terms.

Its role is to model methylation change across repeated observations:

- Main inputs: combined longitudinal phenotype-methylation data, participant
  identifiers, timepoint variables, phenotypes, covariates, and annotation
  settings.
- Main processing tasks: longitudinal model preparation, CpG-LME model
  fitting, interaction testing, diagnostic plotting, summarisation, and
  annotation.
- Main outputs: mixed-model objects, interaction summaries, diagnostic figures,
  significant-interaction exports, and annotated longitudinal workbooks.

```{r methylationGLMM_T1T2-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["methylationGLMM_T1T2"]])
```

## dnamReport()

`dnamReport()` assembles the main tables, figures, logs, and model summaries
into a report website. It is usually run after preprocessing and modelling
outputs have been written to disk.

Its role is to make the pipeline outputs easier to inspect and share:

- Main inputs: phenotype tables, QC figures, ENmix figures, SVA figures,
  metric figures, model annotation tables, detection P-value data, and logs.
- Main processing tasks: input validation, report file preparation, Quarto
  rendering, and post-processing.
- Main outputs: a rendered report site with organised tabs for data,
  preprocessing, modelling, logs, and supporting diagnostics.

```{r dnamReport-overview, echo = FALSE, results = "asis"}
svg_to_html(
  overview_svgs[["dnamReport"]])
```

# Summary

The `dnaEPICO` pipeline is organised around function functions with clear roles.
The preprocessing functions create quality-controlled methylation data and
analysis-ready phenotype tables. The modelling functions fit cross-sectional and
longitudinal association models. The report function gathers the resulting
tables, figures, and logs into a browsable output.

In practice, the most common pipeline is:

- run `preprocessingMinfiEwasWater()` to prepare methylation objects and QC
  outputs,
- run `svaEnmix()` when control-probe surrogate variables are needed,
- run `preprocessingPheno()` to prepare modelling tables,
- run `methylationGLM_T1()` or `methylationGLMM_T1T2()` for association
  testing, and
- run `dnamReport()` to review the completed outputs.

# Basics

Date the vignette was generated.

```{r, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r, echo = FALSE}
sessionInfo()
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and
in which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve, so it is critical to learn where to ask for help. We would
like to highlight the [Bioconductor support site](https://support.bioconductor.org/)
as the main resource for getting help. Please remember to use the `dnaEPICO`
tag and check the
[older posts](https://support.bioconductor.org/tag/dnaEPICO/). If you want to
receive help, please provide a small reproducible example and your session
information so the source of the problem can be tracked efficiently.
