---
title: "The dnaEPICO User's Guide"
shorttitle: "dnaEPICO guide"
author: 
  - name: Paul Ruiz
    affiliation:
    - Queensland University of Technology
    email: ruizpint@qut.edu.au
  - name: Divya Mehta
    affiliation:
    - Queensland University of Technology
abstract: >
  A comprehensive guide to use the dnaEPICO package. The package supports preprocessing and analysis of Illumina DNA methylation array data (EPICv2, EPIC and 450K), including QC, filtering, SVA, cell-type deconvolution, phenotype integration, modelling, and export-ready files for the DNAmAge Clock Foundation platform.
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
  %\VignetteIndexEntry{Introduction to dnaEPICO}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Introduction

The **dnaEPICO** package provides a structured workflow for preprocessing and analysing Illumina DNA methylation array data. It works with EPICv2, EPIC, and 450K arrays. The package supports data import, quality control, filtering, normalisation, surrogate variable analysis, cell-type deconvolution, phenotype integration, statistical modelling, and export files for DNAmAge Clock Foundation.

The package is designed to integrate with established Bioconductor tools such as `minfi`, `ENmix`, and `wateRmelon`.

`r Biocpkg("dnaEPICO")` also prepares standardised DNA methylation files for direct export to the [DNAmAge Clock Foundation platform](https://dnamage.clockfoundation.org/user).

## Citing `dnaEPICO`

The `r Biocpkg("dnaEPICO")` package uses methods from several external packages. Because no single manuscript describes all components, the guidance below explains how to cite dnaEPICO depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of [minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd), [ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and [wateRmelon](https://github.com/schalkwyk/wateRmelon/blob/master/vignettes/references.bib).

- If you use `preprocessingMinfiEwasWater()`, please cite @minfi; @minfiEPIC; @ENmix; @wateRmelon; @ewastools; @SWAN; @funnorm; @noob; @quantile.
- If you use `svaEnmix()`, please cite @minfi; @ENmix.
- If you use `make f3 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools.
- If you use `make f4 MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2.
- If you use `make f3lme MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @lme4.
- If you use `make all MODEL=model1`, please cite @minfi; @minfiEPIC; @ENmix; @wateRmelon; @SWAN; @funnorm; @noob; @quantile; @ewastools; @glm2; @lme4.

# Installation

The `r Biocpkg("dnaEPICO")` package depends on `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`, among others. Install dependencies with BiocManager:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("dnaEPICO")

BiocManager::valid()
```

```{r "start", message=FALSE, eval = FALSE}
library("dnaEPICO")
```

# Overview

`r Biocpkg("dnaEPICO")` is a DNA methylation preprocessing and modelling pipeline for Illumina arrays. It wraps standard analysis steps into a consistent structure that:

- reads **IDATs** and phenotype metadata,
- performs **quality control**, filtering, and normalisation,
- computes surrogate variables (**SVA**),
- merges phenotype data with methylation matrices,
- fits **GLM** (cross-sectional) and **GLMM/LME** (longitudinal) models,
- returns structured in-memory result objects for interactive use,
- optionally writes **logs**, **figures**, legacy **RData objects**, and a **PDF report** when requested.

Each step can be run interactively without writing files, or connected through the Makefile pipeline when you set `saveOutputs = TRUE`. This design lets you inspect results in memory first and still reproduce the classic folder-based workflow when needed.

## Local/Cluster use

**Local use** is recommended for:

- testing the pipeline on small data,
- preparing clean merged data objects,
- generating QC plots and logs,
- validating input formats and parameters.

**Cluster/HPC use** is recommended for:

- genome-wide modelling across many CpGs,
- parallel chunk processing,
- large memory and multi-core work,
- reproducible batch runs with Make targets (`f3`, `f4`, `f3lme`, `all`).

# Data expectations and folder conventions

`r Biocpkg("dnaEPICO")` expects a project layout like this:

``` make
├── data/                                   # Processed data outputs (per model)
│   ├── preprocessingMinfiEwasWater/
│   │   ├── idats/                          # Folder containing raw IDAT files
│   │   ├── pheno.csv                       # Phenotype file with sample metadata
│   │   └── 12864_2024_10027_MOESM8_ESM.csv # Cross-reactivity comparison reference
│   ├── Makefile                            # Arguments
```

- **Phenotype file**: a `CSV` with sample identifiers and covariates.
- **IDAT folder**: contains Illumina `*_Red.idat` and `*_Grn.idat` pairs.

Cross-hybridisation probe data for EPICv2 can be downloaded from [`12864_2024_10027_MOESM8_ESM.csv`](https://springernature.figshare.com/articles/dataset/Additional_file_8_of_Characterisation_and_reproducibility_of_the_HumanMethylationEPIC_v2_0_BeadChip_for_DNA_methylation_profiling/26689706?file=48531323).

You can extract the example Makefile with:

```{r, eval=FALSE}
makefilePath <- extractMake(
    destDir = getwd(),
    overwrite = FALSE
)
```

When disk outputs are enabled, `r Biocpkg("dnaEPICO")` produces:

``` make
├── data/                                   # Processed data outputs (per model)
│   ├── preprocessingMinfiEwasWater/
│   │   ├── idats/                          # Folder containing raw IDAT files
│   │   ├── pheno.csv                       # Phenotype file with sample metadata
│   │   └── 12864_2024_10027_MOESM8_ESM.csv # Cross-reactivity comparison reference
│   ├── model1/                             # Outputs specific to the first model
│   │   ├── preprocessingMinfiEwasWater/    # Preprocessed phenotype + QC from Minfi
│   │   ├── svaEnmix/                       # Surrogate variable analysis 
│   │   ├── preprocessingPheno/             # Cleaned and merged phenotype files 
│   │   ├── methylationGLM_T1/              # CpG-level GLM results (single timepoint)
│   │   ├── methylationGLMM_T1T2/           # Longitudinal LME results (T1 vs T2)
│   │   └── ...                             # Additional subfolders by model
│   ├── model2/
│   └── model3/
├── rData/                                  # Processed R objects (MSet, Beta, CN matrices, etc.)
├── logs/                                   # Logging output from script runs
├── figures/                                # QC plots and visualization output
├── results/                                # Output tables and summary statistics
├── reports/                                # Rendered reports (PDF)
```

When `saveOutputs = TRUE`, the workflow writes:

- `logs/`: logs
- `rData/`: intermediate and final R objects
- `figures/`: plots organised by step
- `data/`: exported tables and annotated results
- `reports/`: final report files

This approach improves reproducibility, debugging, and HPC resuming while keeping the functions usable in the traditional in-memory Bioconductor style.

# Local use

Local use focuses on the first three steps that are usually run interactively:

1.  `preprocessingMinfiEwasWater()`: import, QC, normalise, filter, and estimate cell composition
2.  `svaEnmix()`: surrogate variable analysis using ENmix control probes
3.  `preprocessingPheno()`: merge phenotype metadata with methylation matrices

## Step 1: `preprocessingMinfiEwasWater()`

```{r, eval=FALSE}
preprocessResult <- preprocessingMinfiEwasWater(
    phenoFile = "data/preprocessingMinfiEwasWater/pheno.csv",
    idatFolder = "data/preprocessingMinfiEwasWater/idats",
    outputLogs = "logs",
    nSamples = NA,
    SampleID = "Sample_Name",
    arrayType = "IlluminaHumanMethylationEPICv2",
    annotationVersion = "20a1.hg38",
    scriptLabel = "preprocessingMinfiEwasWater",
    baseDataFolder = "rData",
    sepType = "",
    tiffWidth = 2000,
    tiffHeight = 1000,
    tiffRes = 150,
    qcCutoff = 10.5,
    detPtype = "m+u",
    detPThreshold = 0.05,
    normMethods = "adjustedfunnorm",
    sexColumn = "Sex",
    pvalThreshold = 0.01,
    chrToRemove = "chrX,chrY",
    snpsToRemove = "SBE,CpG",
    mafThreshold = 0.1,
    crossReactivePath =
        "data/preprocessingMinfiEwasWater/12864_2024_10027_MOESM8_ESM.csv",
    plotGroupVar = "Sex",
    lcRef = "salivaEPIC",
    phenoOrder = "Sample_Name;Timepoint;Sex;PredSex;Basename;Sentrix_ID;Sentrix_Position",
    lcPhenoDir = "data/preprocessingMinfiEwasWater",
    display = FALSE,
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = FALSE
)
```

The example above shows the interactive, in-memory style. For the folder-based
pipeline and Makefile workflow, set `saveOutputs = TRUE` so later steps can
reuse the saved files.

### What it does:

This wrapper coordinates the modular preprocessing helpers and returns a structured result object. It:

1.  Reads the phenotype CSV (`phenoFile`) and IDAT files (`idatFolder`).
2.  Loads an `RGChannelSet` and performs initial QC.
3.  Builds `MSet`, `RatioSet`, and `GenomicRatioSet` objects.
4.  Computes detection p-values and filters failed probes and samples.
5.  Applies normalisation using `normMethods`.
6.  Filters probes based on detection p-values, sex chromosomes, SNPs, and cross-reactive probes.
7.  Returns the filtered phenotype table, preprocessing objects, metrics, and cell composition estimates in memory.
8.  Optionally writes QC plots, logs, legacy `.RData` outputs, and `phenoLC.csv` when `saveOutputs = TRUE`.

### Parameter guide:

-   `phenoFile`: path to sample metadata CSV.
-   `idatFolder`: folder with raw IDAT pairs.
-   `nSamples`: subset size for local testing (use `NA` to process all).
-   `SampleID`: sample identifier column in the phenotype file.
-   `arrayType`, `annotationVersion`: ensure the correct annotation and manifest.
-   `qcCutoff`, `detPtype`, `detPThreshold`: QC thresholds for probe/sample filtering.
-   `normMethods`: chosen normalisation method.
-   `pvalThreshold`, `chrToRemove`, `snpsToRemove`, `mafThreshold`: probe filtering rules.
-   `crossReactivePath`: path to cross-reactive probe list.
-   `phenoOrder`: order of columns in the final phenotype table.
-   `lcRef`, `lcPhenoDir`: cell type reference and output directory for cell fraction results.
-   `display`: draw exploratory plots on the active device.
-   `verbose`, `logs`: control message output and optional log files.
-   `saveOutputs`: write the classic folder-based outputs instead of relying only on the returned object.

## Step 2: `svaEnmix()`

```{r, eval=FALSE}
svaResult <- svaEnmix(
    phenoFile = "data/preprocessingMinfiEwasWater/phenoLC.csv",
    rgsetData = "rData/preprocessingMinfiEwasWater/objects/RGSet.RData",
    sepType = "",
    outputLogs = "logs",
    nSamples = NA,
    SampleID = "Sample_Name",
    arrayType = "IlluminaHumanMethylationEPICv2",
    annotationVersion = "20a1.hg38",
    SentrixIDColumn = "Sentrix_ID",
    SentrixPositionColumn = "Sentrix_Position",
    ctrlSvaPercVar = 0.90,
    ctrlSvaFlag = 1,
    scriptLabel = "svaEnmix",
    tiffWidth = 2000,
    tiffHeight = 1000,
    tiffRes = 150,
    display = FALSE,
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = TRUE
)
```

### What it does:

This wrapper coordinates the modular `svaEnmix()` helpers. It:

1.  Loads the `RGSet` object from `rgsetData` and phenotype data from `phenoFile`.
2.  Uses ENmix control probes to derive control-based surrogate variables.
3.  Selects the number of surrogate components needed to explain `ctrlSvaPercVar` of variance.
4.  Returns the surrogate-variable results, merged phenotype table, and model summaries in memory.
5.  Optionally writes summary outputs, plots, and updated objects when `saveOutputs = TRUE`.

### Parameter guide:

-   `rgsetData`: path to the saved RGSet object from Step 1.
-   `phenoFile`: phenotype file that includes cell fractions.
-   `ctrlSvaPercVar`: proportion of variance explained by control-derived SVA components.
-   `SentrixIDColumn`, `SentrixPositionColumn`: columns that identify array positions.
-   `display`: draw SVA plots on the active device.
-   `verbose`, `logs`: control message output and optional log files.
-   `saveOutputs`: write exported SVA files for downstream pipeline steps.

## Step 3: `preprocessingPheno()`

```{r, eval=FALSE}
phenoResult <- preprocessingPheno(
    phenoFile = "data/preprocessingMinfiEwasWater/phenoLC.csv",
    sepType = "",
    betaPath =
        "rData/preprocessingMinfiEwasWater/metrics/beta_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData",
    mPath =
        "rData/preprocessingMinfiEwasWater/metrics/m_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData",
    cnPath =
        "rData/preprocessingMinfiEwasWater/metrics/cn_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData",
    SampleID = "Sample_Name",
    timeVar = "Timepoint",
    timepoints = "1,2",
    combineTimepoints = "1,2",
    outputPheno = "data/preprocessingPheno",
    outputRData = "rData/preprocessingPheno/metrics",
    outputRDataMerge = "rData/preprocessingPheno/mergeData",
    sexColumn = "Sex",
    outputLogs = "logs",
    outputDir = "data/preprocessingPheno",
    verbose = FALSE,
    logs = FALSE,
    saveOutputs = TRUE
)
```

### What it does:

This wrapper coordinates the modular `preprocessingPheno()` helpers. It:

1.  Loads phenotype data and methylation matrices (Beta, M, CN).
2.  Aligns samples using `SampleID`.
3.  Creates timepoint-specific subsets based on `timepoints`.
4.  Combines timepoints for longitudinal analysis using `combineTimepoints`.
5.  Returns aligned phenotype-plus-methylation objects for downstream modelling.
6.  Optionally writes cleaned phenotype tables to `outputPheno`, metric objects to `outputRData`, and merged objects to `outputRDataMerge` when `saveOutputs = TRUE`.

### Parameter guide:

-   `phenoFile`: the phenotype file generated in Step 1.
-   `betaPath`, `mPath`, `cnPath`: paths to saved methylation matrices.
-   `SampleID`: sample identifier column used to match data.
-   `timeVar`: column used to define timepoints.
-   `timepoints`: comma-separated list of timepoints to export separately.
-   `combineTimepoints`: list of timepoints to combine for longitudinal analysis.
-   `outputPheno`: directory for cleaned phenotype tables.
-   `outputRData`: directory for timepoint-specific methylation objects.
-   `outputRDataMerge`: directory for merged phenotype-methylation objects.
-   `sexColumn`: column used for biological sex.
-   `outputLogs`: directory for logs.
-   `outputDir`: base directory for exported files.
-   `verbose`, `logs`: control message output and optional log files.
-   `saveOutputs`: write pipeline-ready CSV, ZIP, and `.RData` outputs.

This step produces analysis-ready phenotype-methylation objects for GLM and LME modelling and creates files compatible with the DNAmAge Clock Foundation platform.

# Cluster / HPC use

For full datasets, model fitting across hundreds of thousands of CpGs usually requires:

- large memory,
- parallel chunking,
- long runtimes.

`dnaEPICO` supports this through a Makefile-driven pipeline that can run the whole workflow or selected targets.

## Step 4: `methylationGLM_T1()`

This step is typically executed on **HPC systems** because it is computationally intensive.

```{r, eval=FALSE}
methylationGLM_T1(
    inputPheno = "rData/preprocessingPheno/mergeData/phenoBetaT1.RData",
    outputLogs = "logs",
    outputRData = "rData/methylationGLM_T1/models",
    outputPlots = "figures/methylationGLM_T1",
    phenotypes = "Pheno1,Pheno2,Pheno3,Pheno4",
    covariates = "Sex,Age,Ethnicity,TraumaDefinition,Leukocytes,Epithelial.cells",
    factorVars = "Sex,Ethnicity,TraumaDefinition",
    cpgPrefix = "cg",
    cpgLimit = NA,
    nCores = 32,
    plotWidth = 2000,
    plotHeight = 1000,
    plotDPI = 150,
    interactionTerm = NULL,
    libPath = NULL,
    glmLibs = "glm2",
    prsMap = "Pheno1:PRS_1,Pheno2:PRS_2,Pheno3:PRS_3,Pheno4:PRS_4",
    summaryPval = NA,
    summaryResidualSD = TRUE,
    saveSignificantCpGs = FALSE,
    significantCpGDir = "preliminaryResults/cpgs/methylationGLM_T1",
    significantCpGPval = 0.05,
    saveTxtSummaries = TRUE,
    chunkSize = 10000,
    summaryTxtDir = "preliminaryResults/summary/methylationGLM_T1/glm",
    fdrThreshold = 0.05,
    padjmethod = "fdr",
    annotationPackage = "IlluminaHumanMethylationEPICv2anno.20a1.hg38",
    annotationCols = "Name,chr,pos,UCSC_RefGene_Group,UCSC_RefGene_Name,
                      Relation_to_Island,GencodeV41_Group",
    annotatedGLMOut = "data/methylationGLM_T1",
    logs = TRUE,
    saveOutputs = TRUE
)
```

### What it does:

This function returns a structured in-memory analysis object. It:

1.  Loads a merged phenotype-beta object from `inputPheno`.
2.  Processes CpGs in chunks and fits a [generalised linear model](https://cran.r-project.org/web/packages/glm2/index.html) for each CpG.
3.  Fits one model per phenotype and adjusts for covariates.
4.  Handles interaction terms and PRS mappings when supplied.
5.  Applies multiple-testing correction.
6.  Optionally writes serialized model objects, plots, logs, summaries, and annotated results when `saveOutputs = TRUE`.

### Parameter guide:

-   `inputPheno`: merged phenotype-beta `.RData` input.
-   `outputLogs`: directory for log files.
-   `outputRData`: directory for model objects.
-   `outputPlots`: directory for diagnostic and summary plots.
-   `phenotypes`: comma-separated phenotype variables.
-   `covariates`: covariates included in each model.
-   `factorVars`: categorical covariates.
-   `cpgPrefix`: CpG identifier prefix.
-   `cpgLimit`: number of CpGs to process; `NA` processes all.
-   `nCores`: CPU cores for parallel processing.
-   `plotWidth`, `plotHeight`, `plotDPI`: plot size and resolution.
-   `interactionTerm`: optional interaction term.
-   `libPath`: optional HPC library path.
-   `glmLibs`: GLM package to use.
-   `prsMap`: phenotype-to-PRS mapping.
-   `summaryPval`: threshold for summary outputs.
-   `summaryResidualSD`: include residual SD summaries.
-   `saveSignificantCpGs`: save CpGs passing the significance threshold.
-   `significantCpGDir`: directory for significant CpG tables.
-   `significantCpGPval`: significance threshold for CpGs.
-   `saveTxtSummaries`: save text summaries.
-   `chunkSize`: CpGs per chunk.
-   `summaryTxtDir`: directory for text-based summaries.
-   `fdrThreshold`: FDR threshold.
-   `padjmethod`: adjustment method.
-   `annotationPackage`: annotation package for CpGs.
-   `annotationCols`: annotation fields to include.
-   `annotatedGLMOut`: directory for annotated results.
-   `display`: draw exploratory and diagnostic plots on the active device.
-   `verbose`: print progress messages with `message()`.
-   `logs`: write the same progress messages to `outputLogs`.
-   `saveOutputs`: write optional files to disk; otherwise return results in memory only.

### GLM structure used in dnaEPICO (`glm2`):

For each CpG site, dnaEPICO fits a model of the form:

$$
\text{Methylation}_{CpG_i} =
\beta_0
+
\beta_1 \cdot \text{Phenotype}
+
\sum_{k=1}^{K} \beta_{k+1} \cdot \text{Covariate}_k
+
\varepsilon
$$

where $\text{Methylation}_{CpG_i}$ is the methylation value at CpG site $i$, $\beta_0$ is the intercept, $\text{Phenotype}$ is the variable of interest, $\text{Covariate}_k$ represents the $k$-th covariate in the model, $\beta_1$ and $\beta_{k+1}$ are the corresponding regression coefficients, and $\varepsilon$ is the residual error term.

For example, a GLM without PRS is:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} = \;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot \text{Age} +
\beta_3 \cdot \text{Sex} +
\beta_4 \cdot \text{Ethnicity} \\
&+
\beta_5 \cdot \text{TraumaDefinition} +
\beta_6 \cdot \text{Leukocytes} +
\beta_7 \cdot \text{Epithelial.cells} +
\varepsilon
\end{aligned}
$$

### GLM with PRS:

When `prsMap` is supplied, the model is extended only for the matching phenotype.

From the example above:

**prsMap = "Pheno1:PRS_1,Pheno2:PRS_2,Pheno3:PRS_3,Pheno4:PRS_4"**

This defines one-to-one mappings between phenotype and PRS. For example:

-   Pheno1 uses PRS_1

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Pheno1} +
\sum_{k=1}^{K} \beta_{k+1} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS_1} +
\varepsilon
$$

-   Pheno2 uses PRS_2

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot \text{Pheno2} +
\sum_{k=1}^{K} \beta_{k+1} \cdot \text{Covariate}_k +
\beta_{\text{PRS}} \cdot \text{PRS_2} +
\varepsilon
$$

### Interaction term

If an interaction is specified (for example phenotype Ã— timepoint):

$$
\text{Methylation}_{CpG_i} =
\beta_0 +
\beta_1 \cdot (\text{Phenotype} * \text{Timepoint}) +
\sum_{k=1}^{K} \beta_{k+1} \cdot \text{Covariate}_k +
\varepsilon
$$

## Step 5: `methylationGLMM_T1T2()`

This step is typically executed on **HPC systems** because it is also computationally intensive.

```{r, eval=FALSE}
methylationGLMM_T1T2(
    inputPheno = "rData/preprocessingPheno/mergeData/phenoBetaT1T2.RData",
    outputLogs = "logs",
    outputRData = "rData/methylationGLMM_T1T2/models",
    outputPlots = "figures/methylationGLMM_T1T2",
    personVar = "person",
    timeVar = "Timepoint",
    phenotypes = "Pheno1,Pheno2,Pheno3,Pheno4",
    covariates = "Sex,Age,Ethnicity,TraumaDefinition,Leukocytes,Epithelial.cells",
    factorVars = "Sex,Ethnicity,TraumaDefinition,Timepoint",
    lmeLibs = "lme4,lmerTest",
    prsMap = "Pheno1:PRS_1,Pheno2:PRS_2,Pheno3:PRS_3,Pheno4:PRS_4",
    libPath = NULL,
    cpgPrefix = "cg",
    cpgLimit = NA,
    nCores = 32,
    summaryPval = NA,
    plotWidth = 2000,
    plotHeight = 1000,
    plotDPI = 150,
    interactionTerm = NULL,
    saveSignificantInteractions = TRUE,
    significantInteractionDir = "preliminaryResults/cpgs/methylationGLMM_T1T2",
    significantInteractionPval = 0.05,
    saveTxtSummaries = TRUE,
    chunkSize = 10000,
    summaryTxtDir = "preliminaryResults/summary/methylationGLMM_T1T2/lmer",
    fdrThreshold = 0.05,
    padjmethod = "fdr",
    annotationPackage = "IlluminaHumanMethylationEPICv2anno.20a1.hg38",
    annotationCols = "Name,chr,pos,UCSC_RefGene_Group,UCSC_RefGene_Name,
                      Relation_to_Island,GencodeV41_Group",
    annotatedLMEOut = "data/methylationGLMM_T1T2",
    display = FALSE,
    verbose = FALSE,
    logs = TRUE,
    saveOutputs = TRUE
)
```

### What it does:

This function returns a structured in-memory longitudinal methylation analysis object. It:

1.  Loads a merged longitudinal phenotype-methylation object from `inputPheno`.
2.  Processes CpGs in chunks and fits a linear mixed-effects model.
3.  Fits one model per phenotype and uses a subject-level random effect.
4.  Includes fixed effects for phenotype, time, covariates, and optional interactions.
5.  Adds phenotype-specific PRS terms based on `prsMap`.
6.  Applies multiple-testing correction.
7.  Optionally writes model objects, figures, logs, summaries, significant interaction results, and annotated tables when `saveOutputs = TRUE`.

### Parameter guide:

-   `inputPheno`: merged longitudinal phenotype-methylation `.RData` or `.rds` input.
-   `outputLogs`: directory for logs.
-   `outputRData`: directory for mixed-effects model objects.
-   `outputPlots`: directory for plots.
-   `personVar`: participant identifier for the random effect.
-   `timeVar`: time variable for longitudinal analysis.
-   `phenotypes`: comma-separated phenotype variables.
-   `covariates`: fixed-effect covariates.
-   `factorVars`: categorical variables including `timeVar`.
-   `lmeLibs`: mixed-model libraries.
-   `prsMap`: phenotype-to-PRS mappings.
-   `libPath`: optional library path.
-   `cpgPrefix`: CpG identifier prefix.
-   `cpgLimit`: maximum CpGs; `NA` uses all.
-   `nCores`: parallel cores.
-   `summaryPval`: summary threshold; `NA` disables filtering.
-   `plotWidth`, `plotHeight`, `plotDPI`: plot dimensions and resolution.
-   `interactionTerm`: optional interaction term.
-   `saveSignificantInteractions`: save significant interactions.
-   `significantInteractionDir`: directory for interaction results.
-   `significantInteractionPval`: significance threshold for interactions.
-   `saveTxtSummaries`: save text summaries.
-   `chunkSize`: CpGs per chunk.
-   `summaryTxtDir`: directory for text summaries.
-   `fdrThreshold`: FDR threshold.
-   `padjmethod`: adjustment method.
-   `annotationPackage`: annotation package.
-   `annotationCols`: annotation fields.
-   `annotatedLMEOut`: directory for annotated results.
-   `display`: draw diagnostic plots on the active device.
-   `verbose`: emit progress messages with `message()`.
-   `logs`: save the same progress messages in `outputLogs`.
-   `saveOutputs`: write files to the requested output folders; otherwise results stay in memory.

### LMEM structure used in dnaEPICO (`lme4`):

For longitudinal analyses, the model includes random effects and optional interactions:

$$
\begin{aligned}
\text{Methylation}_{CpG_i} = \;&
\beta_0 +
\beta_1 \cdot (\text{Phenotype} * \text{Timepoint}) +
\beta_2 \cdot \text{covariate} +
\beta_3 \cdot \text{PRS} +
(1 | person)+
\varepsilon
\end{aligned}
$$

where `PRS` is included only if it is defined for the phenotype.

The model follows the same general form as `methylationGLM_T1()`.

$$
\begin{aligned}
\text{Methylation}_{CpG_i} =\;&
\beta_0 +
\beta_1 \cdot \text{Phenotype} +
\beta_2 \cdot \text{Timepoint} +
\beta_3 \cdot (\text{Phenotype} \times \text{Timepoint}) \\
&+
\beta_4 \cdot \text{Age} +
\beta_5 \cdot \text{PRS} +
 b_{\text{person}} +
\varepsilon
\end{aligned}
$$

where $b_{person}$ is the subject-specific random intercept.

# Makefile pipeline concepts

-   `Makefile.model.pipeline`: user configuration (models, variables, paths)
-   `Makefile.rules.pipeline`: pipeline targets and rules

Extract the example Makefile into the current directory:

```{r, eval=FALSE}
## Copies the example Makefile into the current directory
makefilePath <- extractMake(destDir = getwd(), overwrite = FALSE)
```

The example template `Makefile.model.pipeline` is copied and renamed to `Makefile`.

### Make targets: f3, f4, f3lme and all

Common grouped targets:

-   `f3`: Steps 1-3 (preprocessing + SVA + merge)
-   `f4`: Steps 1-4 (adds GLM)
-   `f3lme`: Steps 1-3 and 5 (adds LME4)
-   `all`: Steps 1-5

### Commands

-   Show status of pipeline outputs
    -   make status MODEL=model1
-   Run steps 1-3 for one model
    -   make f3 MODEL=model1
-   Run steps 1-4 for one model
    -   make f4 MODEL=model1
-   Run steps 1-3 and 5 for one model
    -   make f3lme MODEL=model1
-   Run the full pipeline
    -   make all MODEL=model1
-   Run multiple models
    -   make models *(run the entire pipeline for all models in parallel.)*
    -   make f3_models *(run Steps 1-3 in parallel for all models.)*
    -   make f4_models *(run Steps 1-4 in parallel for all models.)*
    -   make f3lme_models *(run Steps 1-3 + LME in parallel for all models.)*

## Makefile Modelling steps (HPC)

The Makefile controls how dnaEPICO runs models at scale. It defines:

-   which phenotype files to use,
-   which statistical models to run,
-   GLM (`glm2`) and GLMM (`lme4`) parameter settings,
-   how results are organised and reproduced across models.

Instead of hard-coding parameters in R scripts, dnaEPICO uses a declarative Makefile layer. This makes it easy to rerun the same workflow with different configurations.

``` make
# ===============================================
# USER CONFIGURATION
# ===============================================
MAKEFLAGS += --output-sync

# ===============================================
# MODELS SELECTION PARALLEL RUN
# ===============================================
MODEL ?= model1 # Single model
MODELS = modelA modelB modelC # Multiple models for parallel run

# ===============================================
# PER-MODEL OVERRIDES
# ===============================================
ifeq ($(MODEL), modelA)
  PHENO_FILE = $(DATA_DIR)/$(STEP1)/phenoA.csv


else ifeq ($(MODEL), modelB)
  PHENO_FILE = $(DATA_DIR)/$(STEP1)/phenoB.csv


else ifeq ($(MODEL), modelC)
  PHENO_FILE = $(DATA_DIR)/$(STEP1)/phenoC.csv

else # default (model1)
  PHENO_FILE = $(DATA_DIR)/$(STEP1)/pheno.csv

endif

# ===============================================
# DIRECTORIES
# ===============================================
LOGS_DIR = logs
DATA_DIR = data
RDATA_DIR = rData
RESULTS_DIR = results
FIGURES_DIR = figures
STEP1 = preprocessingMinfiEwasWater
STEP2 = svaEnmix
STEP3 = preprocessingPheno
STEP4 = methylationGLM_T1
STEP5 = methylationGLMM_T1T2
STEP6 = reports
METRICS_DIR = metrics
IDAT_DIR = idats
OBJ_DIR = objects
MERGE_DIR = mergeData
MODEL_DIR = models
CPG_DIR = cpgs
SUMMARY_DIR = summary
ENMIX_DIR = enmix
SVA_DIR = sva
QC_DIR = qc
R_DIR = ~/R/x86_64-pc-linux-gnu-library/4.4 # NULL is default R library path

# ===============================================
# GLOBAL PARAMETERS
# ===============================================
# STEP 1-5 PARAMETERS
PHENO_FILE = $(DATA_DIR)/$(STEP1)/phenoEMD.csv
SEP_TYPE = ""
SAMPLE_ID = Sample_Name
N_SAMPLES = 30
ARRAY_TYPE = IlluminaHumanMethylationEPICv2
ANNOTATION_VERSION = 20a1.hg38
TIFF_WIDTH = 2000
TIFF_HEIGHT = 1000
TIFF_RES = 150
SEX_COLUMN = Sex
SENTRIX_ID_COLUMN = Sentrix_ID
SENTRIX_POSITION_COLUMN = Sentrix_Position
BASENAME_COLUMN = Basename
TIME_VAR = Timepoint
PHENO_ORDER = $(SAMPLE_ID);$(TIME_VAR);$(SEX_COLUMN);PredSex;$(BASENAME_COLUMN);$(SENTRIX_ID_COLUMN);$(SENTRIX_POSITION_COLUMN)

# STEP 4-5 PARAMETERS
PHENOTYPES = TreatmentGroup
COVARIATES = Sex
FACTOR_VARS = Sex,TreatmentGroup
CPG_PREFIX = cg
CPG_LIMIT = NA
PRS_MAP = NULL
SUMMARY_PVAL = NA
N_CORES = 64
SAVE_TXT_SUMMARIES = TRUE
CHUNK_SIZE = 10000
FDR_THRESHOLD = 0.05
```

### Model selection

-   MODEL ?= model1
-   MODELS = modelA modelB modelC
-   `MODEL` defines **a single pipeline configuration**
-   `MODELS` defines **multiple independent configurations**

The same pipeline can therefore be run in two modes:

- **Single-model execution**
  - make f4 MODEL=modelA

- **Parallel multi-model execution**
  - make f4_models

In parallel mode, **each model runs as a fully independent pipeline**, with its own inputs, logs, figures, and outputs. No files are shared between models except the raw inputs (IDATs), unless explicitly specified.

## Per-model overrides

``` make
ifeq ($(MODEL), modelA)
  PHENO_FILE = data/preprocessingMinfiEwasWater/phenoA.csv
...
else
  PHENO_FILE = data/preprocessingMinfiEwasWater/pheno.csv
endif
```

This block **dynamically replaces variables** based on the selected model.

-   The Makefile is evaluated **at runtime**
-   When `MODEL=modelA`, `PHENO_FILE` is replaced before any step runs
-   The rest of the pipeline remains unchanged

This allows users to:

-   compare cohorts or subgroup definitions,
-   test alternative phenotype encodings,
-   run sensitivity analyses, without duplicating code.

### Directory and step conventions

-   STEP1 = preprocessingMinfiEwasWater
-   STEP2 = svaEnmix
-   STEP3 = preprocessingPheno
-   STEP4 = methylationGLM_T1
-   STEP5 = methylationGLMM_T1T2

In the Makefile pipeline, each step writes outputs into **step-specific
directories**. When running in parallel, directory paths are automatically
namespaced by model, ensuring isolation between runs.

This design allows:

-   restarting individual steps,
-   debugging failed models independently,
-   mixing local and HPC execution safely.

## Global parameters

-   PHENO_FILE = ...
-   SAMPLE_ID = ...
-   TIME_VAR = ...
-   ...

These parameters are **shared across preprocessing and modelling**.

### Specific parameters

-   PHENOTYPES = TreatmentGroup
-   COVARIATES = Sex
-   FACTOR_VARS = Sex,TreatmentGroup
-   N_CORES = 64
-   CHUNK_SIZE = 10000
-   ...

These parameters control **how each model is fit**. In parallel execution:

-   `N_CORES` controls **intra-model parallelism** (within R)
-   `MODELS` controls **inter-model parallelism** (within Make)

### Argument normalisation

``` make
ifeq ($(PRS_MAP),NULL)
  PRS_MAP_ARG = NULL
else
  PRS_MAP_ARG = '$(PRS_MAP)'
endif
if ...
```

This section ensures that optional arguments are passed correctly to R scripts. It prevents accidental interpretation of the string `"NULL"` as a real value and guarantees consistent behaviour across local and cluster environments.

# Reporting

The reporting step follows the same modular pattern as the analysis steps:

1.  `prepareDnamReportInputs()` collects report metadata and figure inventories.
2.  `renderDnamReport()` renders a prepared report object.
3.  `dnamReport()` is a convenience wrapper around both steps.

Use the split API when you want to inspect the report inputs before rendering. Use `dnamReport()` when you want the convenience wrapper.

Note: PDF report generation requires a working LaTeX installation. TinyTeX can be installed in R using `install.packages("tinytex")` and `tinytex::install_tinytex()`.
For troubleshooting LaTeX-related issues, see the [TinyTeX](https://yihui.org/tinytex/) documentation.

```{r, eval=FALSE}
preparedReport <- prepareDnamReportInputs(
    output = "DNAm_Report.pdf",
    outputDir = "reports",
    qcDir = "figures/preprocessingMinfiEwasWater/enMix",
    preprocessingDir = "figures/preprocessingMinfiEwasWater/qc",
    postprocessingDir = "figures/preprocessingMinfiEwasWater/metrics",
    svaDir = "figures/svaEnmix/sva",
    glmDir = "figures/methylationGLM_T1",
    glmmDir = "figures/methylationGLMM_T1T2",
    reportTitle = "DNA methylation",
    author = "School of Biomedical Sciences",
    date = format(Sys.Date(), "%B %d, %Y"),
    verbose = FALSE,
    logs = FALSE
)

reportRender <- renderDnamReport(
    preparedReport = preparedReport,
    verbose = FALSE,
    logs = FALSE
)

reportInfo <- dnamReport(
    output = "DNAm_Report.pdf",
    outputDir = "reports",
    qcDir = "figures/preprocessingMinfiEwasWater/enMix",
    preprocessingDir = "figures/preprocessingMinfiEwasWater/qc",
    postprocessingDir = "figures/preprocessingMinfiEwasWater/metrics",
    svaDir = "figures/svaEnmix/sva",
    glmDir = "figures/methylationGLM_T1",
    glmmDir = "figures/methylationGLMM_T1T2",
    reportTitle = "DNA methylation",
    author = "School of Biomedical Sciences",
    date = format(Sys.Date(), "%B %d, %Y"),
    render = TRUE,
    verbose = FALSE,
    logs = FALSE
)
```

# Troubleshooting

- If a step "fails", check logs in `logs/`
- If Make targets are missing, confirm:
  - your working directory contains a valid `Makefile`

# Example Analysis

The following examples are brief demonstrations on how to perform DNA methylation analysis using `dnaEPICO` functions and arguments.

This example has the following dependencies:

```{r dependencies, warning=FALSE, message=FALSE}
library(dnaEPICO)
```

```{r, warning=FALSE, message=FALSE}
library(minfiData)
library(minfi)
```

## Example 1: Local Test

The aim of this example is to show **how to run the complete `dnaEPICO` pipeline** using the example dataset provided by the `minfiData` package.

This example is designed to:

- help new users understand the pipeline structure,
- demonstrate how data and results are organised on disk,
- and provide a safe, reproducible way to test the pipeline

To avoid creating permanent files on the user's computer, **all steps are run inside a temporary directory**.

When analysing real data, users can simply replace the temporary directory with their own project folder.

### Overview of example 1

In this example, the pipeline performs the following steps:

1.  Stage raw input data (IDAT files and phenotype data)
2.  Perform DNA methylation preprocessing and quality control
3.  Estimate surrogate variables (technical variation)
4.  Prepare phenotype-aligned methylation matrices
5.  Generate figures and a summary PDF report

### Step 1: Create a temporary working directory

We first create a temporary directory that will hold **all input files and outputs** for this example.

```{r, message=FALSE, warning=FALSE, include=TRUE}
# Temporary directory
tmpDir <- file.path(tempdir(), "dnaEPICO_Test")
dir.create(tmpDir, showWarnings = FALSE, recursive = TRUE)
```

```{r, message=FALSE, warning=FALSE, include=TRUE}
preprocDir <- file.path(tmpDir, "data", "preprocessingMinfiEwasWater")
dir.create(preprocDir, showWarnings = FALSE, recursive = TRUE)

# Register automatic cleanup when the vignette exits
on.exit(unlink(tmpDir, recursive = TRUE, force = TRUE), add = TRUE)

tmpDir
```

**Important:**

-   This directory is created only for this example
-   It will be deleted automatically at the end
-   Users analysing real data should use a permanent directory instead

### Step 2: Load example IDAT files from minfiData

The `minfiData` package contains a small set of Illumina 450K IDAT files from blood samples.

```{r download-raw, message=FALSE, warning=FALSE }
# Locate minfiData extdata directory
baseDir_minfi <- system.file("extdata", package = "minfiData")
```

```{r, message=FALSE, warning=FALSE }
# Find all IDAT files
idatFiles <- list.files(
    baseDir_minfi,
    pattern = "\\.idat$",
    recursive = TRUE,
    full.names = TRUE
)

length(idatFiles)
```

We now copy these IDAT files into the directory structure expected by `dnaEPICO`.

```{r}
# Create IDAT directory
idatDir <- file.path(preprocDir, "idats")
dir.create(idatDir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
# Copy IDATs into idatDir
file.copy(idatFiles, idatDir, overwrite = TRUE)

length(list.files(idatDir, pattern = "\\.idat$"))
```

### Step 3: Copy the cross-reactive probe annotation file

The pipeline requires a file listing cross-reactive probes. This file is distributed with `dnaEPICO` and is copied into the preprocessing folder. It is a reduced version of the original file, which can be accessed from the following link:

-   [`12864_2024_10027_MOESM8_ESM.csv`](https://springernature.figshare.com/articles/dataset/Additional_file_8_of_Characterisation_and_reproducibility_of_the_HumanMethylationEPIC_v2_0_BeadChip_for_DNA_methylation_profiling/26689706?file=48531323)

```{r}
# Locate dnaEPICO extdata directory
baseDir_dnaEPICO <- system.file("extdata", package = "dnaEPICO")
```

```{r}
# Source cross-reactive probe file
crossReactiveSrc <- file.path(baseDir_dnaEPICO, "12864_2024_10027_MOESM8_ESM.csv")
```

```{r}
# Destination path inside tmpDir/data/preprocessingMinfiEwasWater/
crossReactivePath <- file.path(preprocDir, "12864_2024_10027_MOESM8_ESM.csv")
```

```{r}
# Copy into the target folder
file.copy(crossReactiveSrc, crossReactivePath, overwrite = TRUE)
file.exists(crossReactivePath)
```

### Step 4: Create a phenotype file

Next, we create a phenotype file using the sample sheet provided by `minfiData`.

```{r}
# Read SampleSheet from minfiData
targets <- read.csv(
    file.path(baseDir_minfi, "SampleSheet.csv"),
    stringsAsFactors = FALSE,
    skip = 7
)
```

```{r}
# Create Basename column
targets$Basename <- paste0(
    targets$Sentrix_ID,
    "_",
    targets$Sentrix_Position
)
```

```{r}
# Create Timepoint column
targets$Timepoint <- 1
```

```{r}
# Write phenotype file
samplePath <- file.path(preprocDir, "sample_minfi.csv")
write.csv(targets, samplePath, row.names = FALSE)

file.exists(samplePath)
```

This phenotype file will be used throughout the pipeline.

### Step 5: Preprocessing and quality control

We now run the first pipeline stage, which performs:

-   reading of IDAT files, 
-   quality control, 
-   normalisation, 
-   probe filtering, 
-   and generation of QC figures

```{r}
set.seed(123)
preprocessResult <- preprocessingMinfiEwasWater(
    phenoFile = samplePath,
    idatFolder = file.path(preprocDir, "idats"),
    nSamples = NA,
    SampleID = "Sample_Name",
    phenoOrder = "Sample_Name;sex;Basename;Sentrix_ID;Sentrix_Position",
    arrayType = "IlluminaHumanMethylation450k",
    annotationVersion = "ilmn12.hg19",
    sexColumn = "sex",
    plotGroupVar = "sex",
    tiffWidth = 2000,
    tiffHeight = 1000,
    tiffRes = 150,
    qcCutoff = 10.5,
    detPtype = "m+u",
    detPThreshold = 0.05,
    normMethods = "adjustedfunnorm",
    pvalThreshold = 0.01,
    chrToRemove = "chrX,chrY",
    snpsToRemove = "SBE,CpG",
    mafThreshold = 0.1,
    outputLogs = file.path(tmpDir, "logs"),
    baseDataFolder = file.path(tmpDir, "rData"),
    figureBaseDir = file.path(tmpDir, "figures"),
    lcRef = "FlowSorted.Blood.450k",
    crossReactivePath = file.path(preprocDir, "12864_2024_10027_MOESM8_ESM.csv"),
    lcPhenoDir = preprocDir,
    saveOutputs = TRUE
)

inherits(preprocessResult, "dnaEPICO_preprocessingMinfiEwasWater")
```

### Step 6: Surrogate variable analysis (SVA)

Next, we estimate surrogate variables to capture technical variation such as chip and position effects.

```{r}
rgsetPath <- file.path(
    tmpDir,
    "rData",
    "preprocessingMinfiEwasWater",
    "objects",
    "RGSet.RData"
)

file.exists(rgsetPath)
```

```{r}
phenoLCPath <- file.path(preprocDir, "phenoLC.csv")
file.exists(phenoLCPath)
```

```{r}
svaResult <- svaEnmix(
    phenoFile = phenoLCPath,
    rgsetData = rgsetPath,
    sepType = "",
    outputLogs = file.path(tmpDir, "logs"),
    nSamples = NA,
    SampleID = "Sample_Name",
    arrayType = "IlluminaHumanMethylation450k",
    annotationVersion = "ilmn12.hg19",
    SentrixIDColumn = "Sentrix_ID",
    SentrixPositionColumn = "Sentrix_Position",
    ctrlSvaPercVar = 0.90,
    ctrlSvaFlag = 1,
    scriptLabel = "svaEnmix",
    tiffWidth = 2000,
    tiffHeight = 1000,
    tiffRes = 150,
    dataBaseDir = file.path(tmpDir, "data"),
    figureBaseDir = file.path(tmpDir, "figures"),
    rBaseDir = file.path(tmpDir, "rData"),
    saveOutputs = TRUE
)

inherits(svaResult, "dnaEPICO_svaEnmix")
```

### Step 7: Phenotype preprocessing

This step aligns phenotype data with the processed methylation matrices.

```{r}
# Metric paths produced by preprocessingMinfiEwasWater
betaPath <- file.path(
    tmpDir,
    "rData",
    "preprocessingMinfiEwasWater",
    "metrics",
    "beta_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData"
)
```

```{r}
mPath <- file.path(
    tmpDir,
    "rData",
    "preprocessingMinfiEwasWater",
    "metrics",
    "m_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData"
)
```

```{r}
cnPath <- file.path(
    tmpDir,
    "rData",
    "preprocessingMinfiEwasWater",
    "metrics",
    "cn_NomFilt_MSetF_Flt_Rxy_Ds_Rc.RData"
)
```

```{r}
# Quick sanity check
file.exists(betaPath)
file.exists(mPath)
file.exists(cnPath)
```

```{r}
phenoResult <- preprocessingPheno(
    phenoFile = phenoLCPath,
    sepType = "",
    betaPath = betaPath,
    mPath = mPath,
    cnPath = cnPath,
    SampleID = "Sample_Name",
    timeVar = "Timepoint",
    timepoints = "1",
    combineTimepoints = "1",
    sexColumn = "sex",
    outputPheno = file.path(tmpDir, "data", "preprocessingPheno"),
    outputRData = file.path(tmpDir, "rData", "preprocessingPheno", "metrics"),
    outputRDataMerge = file.path(tmpDir, "rData", "preprocessingPheno", "mergeData"),
    outputLogs = file.path(tmpDir, "logs"),
    outputDir = file.path(tmpDir, "data", "preprocessingPheno"),
    saveOutputs = TRUE
)

inherits(phenoResult, "dnaEPICO_preprocessingPheno")
```

### Step 8: Generate a PDF report

Finally, we prepare a summary report configuration that can be rendered once the
full set of figures is available. This keeps the example in the in-memory-first
style while still showing the report inputs that would be rendered later.

```{r}
preparedReport <- prepareDnamReportInputs(
    output = "DNAm_Report.pdf",
    outputDir = file.path(tmpDir, "reports"),
    qcDir = file.path(
        tmpDir, "figures", "preprocessingMinfiEwasWater", "enMix"
    ),
    preprocessingDir = file.path(
        tmpDir, "figures", "preprocessingMinfiEwasWater", "qc"
    ),
    postprocessingDir = file.path(
        tmpDir, "figures", "preprocessingMinfiEwasWater", "metrics"
    ),
    svaDir = file.path(
        tmpDir, "figures", "svaEnmix"
    ),
    glmDir = file.path(
        tmpDir, "figures", "methylationGLM_T1"
    ),
    glmmDir = file.path(
        tmpDir, "figures", "methylationGLMM_T1T2"
    ),
    figDir = file.path(
        tmpDir, "reports", "figures"
    ),
    reportTitle = "DNA methylation",
    author = "School of Biomedical Sciences",
    date = format(Sys.Date(), "%B %d, %Y"),
    verbose = FALSE,
    logs = FALSE
)

preparedReport$outputFile
```

## Example 2: Cluster Test

For large studies or cluster environments, the same workflow can be executed via a Makefile.

```{r, message=TRUE, warning=FALSE, eval=FALSE}
# Extract Makefile
makefilePath <- extractMake(destDir = tmpDir)

oldWd <- getwd()
setwd(tmpDir)
on.exit(setwd(oldWd), add = TRUE)

system(paste(
    "make f3",
    "MODEL=model1",
    "PHENO_FILE=data/preprocessingMinfiEwasWater/sample_minfi.csv",
    "ARRAY_TYPE=IlluminaHumanMethylation450k",
    "ANNOTATION_VERSION=ilmn12.hg19",
    "SEX_COLUMN=sex",
    "PLOT_GROUP_VAR=sex",
    "LC_REF=FlowSorted.Blood.450k",
    "TIMEPOINTS=1",
    "COMBINE_TIMEPOINTS=1",
    "R_DIR=NULL",
    "N_SAMPLES=NA",
    "N_CORES=1"
))
```

### Step 1: Selecting a single test model

-   MODEL ?= model1

For this example, the pipeline is run using **one explicit model** (`model1`) rather than multiple parallel models. This mirrors the single-dataset workflow demonstrated earlier in R.

### Step 2: Using the example phenotype file

-   PHENO_FILE = (DATA_DIR)/(STEP1)/sample_minfi.csv

The phenotype file generated from `minfiData` (`sample_minfi.csv`) is used for consistency between:

-   interactive R execution
-   Makefile-driven execution

### Step 3: Removing custom R library paths

-   R_DIR = NULL

Setting `R_DIR = NULL` tells the pipeline to use the **default R library paths**, which is appropriate for local testing. You need to use a directory if you are running this in HPC environments.

### Step 4: Matching columns

`GLOBAL PARAMETERS`:

- SAMPLE_ID = Sample_Name 
- ARRAY_TYPE = IlluminaHumanMethylation450k
- ANNOTATION_VERSION = ilmn12.hg19 
- SEX_COLUMN = sex 
- SENTRIX_ID_COLUMN = Sentrix_ID 
- SENTRIX_POSITION_COLUMN = Sentrix_Position

These parameters ensure the Makefile uses the:

-   correct sample identifier,
-   correct array platform and annotation,
-   and correct sex column and chip metadata

### Step 5: Plot grouping and cell composition reference

-   PLOT_GROUP_VAR = sex
-   LC_REF = FlowSorted.Blood.450k

Plots are grouped by `sex`, matching earlier QC figures. `FlowSorted.Blood.450k` is selected as the reference panel because `minfiData` represents blood samples.

### Step 6: Timepoint

-   TIMEPOINTS = 1
-   COMBINE_TIMEPOINTS = 1

The example dataset contains a **single timepoint**, so longitudinal logic is disabled.

# Basics

Date the vignette was generated.

```{r, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r echo=FALSE}
sessionInfo()
```

## Acknowledgement

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.

This vignette was generated using `r Biocpkg("BiocStyle")`, with `r CRANpkg("knitr")` and `r CRANpkg("rmarkdown")` running behind the scenes.

## Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for high-dimensional genomic data, with a focus on Illumina DNA methylation arrays. To read this vignette, we assume that you are already familiar with the general DNA methylation workflow. If not, we recommend first reading this [tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html), which provides a practical introduction to the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`. Downstream statistical modelling relies on base R and CRAN frameworks, including
generalised linear models and linear mixed-effects models. Users are expected to have basic familiarity with R, command-line execution, and Illumina IDAT file structures. Downstream statistical modelling relies on base R and CRAN frameworks, including generalised linear models and linear mixed-effects models. Users are expected to have basic familiarity with R, command-line execution, and Illumina IDAT file structures.

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `dnaEPICO` tag and check [the older posts](https://support.bioconductor.org/tag/dnaEPICO/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

# References


