---
title: "Interactive Local Use of dnaEPICO"
author:
  - name: Paul Ruiz
    affiliation:
      - Queensland University of Technology
    email: ruizpint@qut.edu.au
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
  %\VignetteIndexEntry{Interactive Local Use of dnaEPICO}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL
)

library(dnaEPICO)

can_run_preprocessing <- requireNamespace("minfiData", quietly = TRUE) &&
  requireNamespace("IlluminaHumanMethylation450kmanifest", quietly = TRUE) &&
  requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)

can_run_sva <- requireNamespace("minfiData", quietly = TRUE)

preprocessing_inputs <- NULL
if (isTRUE(can_run_preprocessing)) {
  preprocessing_inputs <- dnaEPICO:::exampleMinfiIdatInputsDnaEpico(n = 6L)
}

sva_targets_file <- NULL
sva_rgset_file <- NULL
if (isTRUE(can_run_sva)) {
  sva_inputs <- dnaEPICO:::exampleMinfiBaseDataDnaEpico()
  sva_temp_dir <- tempdir()
  sva_targets_file <- file.path(sva_temp_dir, "sva_targets.csv")
  sva_rgset_file <- file.path(sva_temp_dir, "sva_RGSet.RData")
  utils::write.csv(sva_inputs$targets, sva_targets_file, row.names = FALSE)
  RGSet <- sva_inputs$RGSet
  save(RGSet, file = sva_rgset_file)
}

pheno_inputs <- dnaEPICO:::examplePreprocessingPhenoStateDnaEpico()
```

# Introduction

This vignette demonstrates the local, interactive use of `dnaEPICO`. In this
mode, the main functions return structured in-memory result objects that can be
inspected directly in the R session. This is the recommended route when you are
checking inputs, reviewing quality-control outputs, or learning the structure of
the package results before moving to larger analyses.

The examples below focus on the first three workflow stages:

1. `preprocessingMinfiEwasWater()`
2. `svaEnmix()`
3. `preprocessingPheno()`

All examples use `saveOutputs = FALSE`, which keeps the workflow in memory and
avoids writing the legacy file outputs.

# Citation

The `r Biocpkg("dnaEPICO")` package uses methods from several external
packages. Because no single manuscript describes all components, the guidance
below explains how to cite `dnaEPICO` depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of
[minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd),
[ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and
[wateRmelon](https://github.com/schalkwyk/wateRmelon/blob/master/vignettes/references.bib).

- If you use `preprocessingMinfiEwasWater()`, please cite @minfi; @minfiEPIC;
  @ENmix; @wateRmelon; @ewastools; @SWAN; @funnorm; @noob; @quantile.
- If you use `svaEnmix()`, please cite @minfi; @ENmix.


# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
To read this vignette, we assume that you are already familiar with the general
DNA methylation workflow. If not, we recommend first reading
[this tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html),
which provides a practical introduction to the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor
tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Users are expected to have basic familiarity with R,
Bioconductor workflows, and Illumina IDAT file structures.

If you are asking yourself the question "Where do I start using
Bioconductor?", you might be interested in
[this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

# Installation

The `r Biocpkg("dnaEPICO")` package depends on `r Biocpkg("minfi")`,
`r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`, among others. Install
dependencies with BiocManager:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("dnaEPICO")

BiocManager::valid()
```

```{r, eval = FALSE, message = FALSE}
library("dnaEPICO")
```

# Step 1: preprocessingMinfiEwasWater

The preprocessing wrapper reads a phenotype table and IDAT files, performs
quality control, normalisation, filtering, and cell-composition estimation, and
returns a list-based result object. The next chunk runs the function on a small
example input and prints the class, element names, and a preview of the
filtered phenotype table.

```{r preprocessing-local, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
preprocessResult <- preprocessingMinfiEwasWater(
  phenoFile = preprocessing_inputs$phenoFile,
  idatFolder = preprocessing_inputs$idatFolder,
  outputLogs = file.path(preprocessing_inputs$tempDir, "logs"),
  nSamples = 6,
  SampleID = "Sample_Name",
  arrayType = preprocessing_inputs$arrayType,
  annotationVersion = preprocessing_inputs$annotationVersion,
  scriptLabel = "preprocessingMinfiEwasWater",
  baseDataFolder = file.path(preprocessing_inputs$tempDir, "rData"),
  figureBaseDir = file.path(preprocessing_inputs$tempDir, "figures"),
  detPThreshold = 1,
  normMethods = "quantile",
  sexColumn = "Sex",
  pvalThreshold = 1,
  chrToRemove = "",
  snpsToRemove = "SBE",
  mafThreshold = 1,
  crossReactivePath = preprocessing_inputs$crossReactivePath,
  plotGroupVar = "Sex",
  lcRef = "saliva",
  phenoOrder = "Sample_Name;Sex;Basename;Sentrix_ID;Sentrix_Position",
  lcPhenoDir = preprocessing_inputs$tempDir,
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(preprocessResult)
names(preprocessResult)
head(preprocessResult$targets[, c(
  "Sample_Name",
  "Sex",
  "Sentrix_ID",
  "Sentrix_Position"
)])
```

The returned object has class `dnaEPICO_preprocessingMinfiEwasWater`. The
`targets` element is the filtered phenotype table aligned to the retained
samples. The remaining elements are nested result objects for the major
preprocessing stages.

Arguments used in this example:

- `phenoFile` and `idatFolder` define the phenotype table and the directory of
  raw IDAT pairs.
- `nSamples = 6` keeps the example small enough for interactive use.
- `SampleID = "Sample_Name"` identifies the column used to align phenotype data
  and methylation objects.
- `arrayType` and `annotationVersion` must match the array platform being
  processed.
- `detPThreshold`, `normMethods`, `pvalThreshold`, `chrToRemove`,
  `snpsToRemove`, and `mafThreshold` control quality filtering and probe
  filtering.
- `crossReactivePath` supplies the cross-reactive probe reference used during
  filtering.
- `sexColumn`, `plotGroupVar`, `lcRef`, and `phenoOrder` control sex
  prediction, quality-control grouping, cell-type estimation, and the structure
  of the returned phenotype table.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
  `saveOutputs = FALSE` keep the example quiet, in memory, and free of file
  outputs.

The next chunk inspects three important elements: the filtered `RGSet`, the
metric matrices, and the phenotype table augmented with cell-composition
estimates.

```{r preprocessing-local-inspect, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
class(preprocessResult$RGSet)
names(preprocessResult$metricsData)
head(preprocessResult$lcData$phenoLC[, 1:6])
```

In local use, these are the most useful objects to inspect first. `RGSet` is
the core methylation object; `metricsData` contains the beta-value, M-value, and
copy-number matrices; and `lcData$phenoLC` is the phenotype table that will be
used by later workflow steps.

# Step 2: svaEnmix

The SVA wrapper loads a phenotype table and a saved `RGSet`, estimates
surrogate variables from ENmix control probes, and returns both the surrogate
variables and the merged phenotype table.

```{r sva-local, eval = can_run_sva, message = FALSE, warning = FALSE}
svaResult <- svaEnmix(
  phenoFile = sva_targets_file,
  rgsetData = sva_rgset_file,
  outputLogs = file.path(tempdir(), "logs"),
  SampleID = "Sample_Name",
  arrayType = "IlluminaHumanMethylation450k",
  annotationVersion = "ilmn12.hg19",
  SentrixIDColumn = "Sentrix_ID",
  SentrixPositionColumn = "Sentrix_Position",
  ctrlSvaPercVar = 0.90,
  ctrlSvaFlag = 1,
  scriptLabel = "svaEnmix",
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(svaResult)
names(svaResult)
dim(svaResult$svaData$sva)
head(svaResult$mergedPheno[, 1:min(6, ncol(svaResult$mergedPheno))])
```

The `svaData$sva` matrix contains one row per sample and one column per
surrogate variable. The `mergedPheno` table is often the most convenient object
to inspect because it combines the original phenotype data and the new SVA
columns in one place.

Arguments used in this example:

- `phenoFile` points to the phenotype table that already includes the cell
  composition estimates from preprocessing.
- `rgsetData` points to the saved `RGSet` used to derive ENmix control-probe
  surrogate variables.
- `SampleID`, `SentrixIDColumn`, and `SentrixPositionColumn` identify the
  sample and array-position columns used in the SVA analysis.
- `ctrlSvaPercVar = 0.90` asks the procedure to keep enough control-derived
  surrogate variables to explain 90% of the control-probe variance.
- `ctrlSvaFlag = 1` enables the control-based SVA workflow.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
  `saveOutputs = FALSE` keep the example interactive and in-memory.

The association analysis between the surrogate variables and array-position
metadata is stored separately in `analysisData`.

```{r sva-local-inspect, eval = can_run_sva, message = FALSE, warning = FALSE}
class(svaResult$analysisData)
names(svaResult$analysisData)
```

This object contains the fitted models and ANOVA summaries used to assess how
the surrogate variables relate to Sentrix chip and position effects.

# Step 3: preprocessingPheno

The phenotype-preparation wrapper aligns the phenotype table with the beta,
M-value, and copy-number matrices; splits the data by timepoint; constructs a
combined longitudinal object; and prepares Clock Foundation export tables.

```{r preprocessingPheno-local, message = FALSE, warning = FALSE}
phenoResult <- preprocessingPheno(
  phenoFile = pheno_inputs$phenoPath,
  betaPath = pheno_inputs$betaPath,
  mPath = pheno_inputs$mPath,
  cnPath = pheno_inputs$cnPath,
  SampleID = "Sample_Name",
  timeVar = "Timepoint",
  timepoints = "1,2",
  combineTimepoints = "1,2",
  outputPheno = file.path(pheno_inputs$tempDir, "data", "preprocessingPheno"),
  outputRData = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "metrics"
  ),
  outputRDataMerge = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "mergeData"
  ),
  sexColumn = "Sex",
  outputLogs = file.path(pheno_inputs$tempDir, "logs"),
  outputDir = file.path(pheno_inputs$tempDir, "clockFoundation"),
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(phenoResult)
names(phenoResult)
names(phenoResult$combinedData)
head(phenoResult$combinedData$phenoBeta[, 1:6])
```

The `combinedData$phenoBeta` table is the key downstream object for modeling.
It combines phenotype columns and CpG beta values in a single data frame that
can be passed to the GLM or GLMM wrappers.

Arguments used in this example:

- `phenoFile`, `betaPath`, `mPath`, and `cnPath` define the phenotype table and
  the three methylation matrices that are aligned in this step.
- `SampleID = "Sample_Name"` is the key used to match samples across the loaded
  objects.
- `timeVar = "Timepoint"`, `timepoints = "1,2"`, and
  `combineTimepoints = "1,2"` define the separate exports and the combined
  longitudinal object.
- `outputPheno`, `outputRData`, `outputRDataMerge`, and `outputDir` still need
  to be specified even when `saveOutputs = FALSE`, because they define the
  canonical output structure for the wrapper.
- `sexColumn = "Sex"` identifies the sex variable used by the helper functions.
- `verbose = FALSE`, `logs = FALSE`, and `saveOutputs = FALSE` keep the example
  focused on the returned object rather than file writing.

The Clock Foundation export objects are returned in the same result, which makes
it possible to inspect the export-ready tables before deciding whether they
should be written to disk.

```{r preprocessingPheno-local-inspect, message = FALSE, warning = FALSE}
names(phenoResult$clockFoundation)
head(phenoResult$clockFoundation$phenoCF)
```

# Summary

In interactive local use, the most important habit is to inspect the returned
objects after each step. In practice, the following checks are usually enough to
confirm that the workflow is behaving as expected:

- `class(result)` to confirm which wrapper produced the object
- `names(result)` to see the main return elements
- `head()` on the phenotype-like tables
- `class()` or `dim()` on the methylation objects and matrices

When you are ready to move from interactive inspection to file-based execution,
the next vignette shows how to use `saveOutputs = TRUE` and the Makefile
workflow. The file-based vignette also describes the regression and
mixed-effects model structures used downstream, including how phenotype-specific
PRS terms are added through `prsMap`.

# Basics

Date the vignette was generated.

```{r, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r, echo = FALSE}
sessionInfo()
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and in
which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve, so it is critical to learn where to ask for help. We would like
to highlight the [Bioconductor support site](https://support.bioconductor.org/)
as the main resource for getting help. Please remember to use the `dnaEPICO`
tag and check the
[older posts](https://support.bioconductor.org/tag/dnaEPICO/). If you want to
receive help, please provide a small reproducible example and your session
information so the source of the problem can be tracked efficiently.
