---
title: "Local Use of dnaEPICO"
author:
  - name: Paul Ruiz
    affiliation:
      - Queensland University of Technology
    email: ruizpint@qut.edu.au
  - name: Divya Mehta
    affiliation:
      - Queensland University of Technology
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
package: dnaEPICO
bibliography: dnaEPICO.bib
vignette: >
  %\VignetteIndexEntry{2. Local Use of dnaEPICO}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
startTime <- Sys.time()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL
)

library(dnaEPICO)

can_run_preprocessing <- requireNamespace("minfiData", quietly = TRUE) &&
  requireNamespace("IlluminaHumanMethylation450kmanifest", quietly = TRUE) &&
  requireNamespace("IlluminaHumanMethylation450kanno.ilmn12.hg19", quietly = TRUE)

can_run_sva <- requireNamespace("minfiData", quietly = TRUE)
```

# Introduction

This vignette demonstrates the local, interactive use of `dnaEPICO`. In this
mode, the main functions return structured in-memory result objects that can be
inspected directly in the R session. This is the recommended route when you are
checking inputs, reviewing quality-control outputs, or learning the structure of
the package results before moving to larger analyses.

The examples below focus on the first three workflow stages:

1. `preprocessingMinfiEwasWater()`
2. `svaEnmix()`
3. `preprocessingPheno()`

All examples use `saveOutputs = FALSE`, which keeps the workflow in memory and
avoids writing the legacy file outputs.

# Citation

The `r Biocpkg("dnaEPICO")` package uses methods from several external
packages. Because no single manuscript describes all components, the guidance
below explains how to cite `dnaEPICO` depending on the functions you use.

This citation guidance is adapted from the vignettes and user guides of
[minfi](https://github.com/hansenlab/minfi/blob/devel/vignettes/minfi.Rmd),
[ENmix](https://github.com/xuz1/ENmix/blob/master/vignettes/ENmix.Rmd), and
[wateRmelon](https://github.com/schalkwyk/wateRmelon/blob/master/vignettes/references.bib).

- If you use `preprocessingMinfiEwasWater()`, please cite @minfi; @minfiEPIC;
  @ENmix; @wateRmelon; @ewastools; @SWAN; @funnorm; @noob; @quantile.
- If you use `svaEnmix()`, please cite @minfi; @ENmix.


# Required knowledge

`r Biocpkg("dnaEPICO")` is built on core Bioconductor infrastructure for
high-dimensional genomic data, with a focus on Illumina DNA methylation arrays.
To read this vignette, we assume that you are already familiar with the general
DNA methylation workflow. If not, we recommend first reading
[this tutorial](https://paulyrp.github.io/2025-cpgpneurogenomics-workshop/tutorial.html),
which provides a practical introduction to the main concepts and analysis steps.

Preprocessing and quality control are performed using established Bioconductor
tools, including `r Biocpkg("minfi")`, `r Biocpkg("ENmix")`, and
`r Biocpkg("wateRmelon")`. Users are expected to have basic familiarity with R,
Bioconductor workflows, and Illumina IDAT file structures.

If you are asking yourself the question "Where do I start using
Bioconductor?", you might be interested in
[this post](https://www.bioconductor.org/install/).

## Cross-reactive probe reference files

The lists of excluded probes depend on the platform. For Illumina
HumanMethylationEPIC v2.0, use the cross-reactive probe file from
[Peters et al. (2024)](https://springernature.figshare.com/articles/dataset/Additional_file_8_of_Characterisation_and_reproducibility_of_the_HumanMethylationEPIC_v2_0_BeadChip_for_DNA_methylation_profiling/26689706?file=48531323);
for Illumina MethylationEPIC (a.k.a. 850k array), use
[Pidsley et al. (2016)](https://springernature.figshare.com/articles/dataset/Additional_file_1_Table_S1_of_Critical_evaluation_of_the_Illumina_MethylationEPIC_BeadChip_microarray_for_whole-genome_DNA_methylation_profiling/4405946?file=7132334);
and for Illumina HumanMethylation450k, use
[Chen et al. (2013)](https://figshare.com/articles/dataset/methylAnalysisDataV3_tar_gz/4800970?file=7896205).

# Installation

The `r Biocpkg("dnaEPICO")` package depends on `r Biocpkg("minfi")`,
`r Biocpkg("ENmix")`, and `r Biocpkg("wateRmelon")`, among others. Install
dependencies with BiocManager:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("dnaEPICO")

BiocManager::valid()
```

```{r, eval = FALSE, message = FALSE}
library("dnaEPICO")
```

# Step 1: preprocessingMinfiEwasWater

The preprocessing function reads a phenotype table and IDAT files, performs
quality control, normalisation, filtering, and cell-composition estimation, and
returns a list-based result object. The next chunk runs the function on a small
example input and prints the class, element names, and a preview of the
filtered phenotype table.

### Create example input files

The next chunk shows how to recreate the temporary files used in this example.
It writes a small phenotype table to a temporary directory, copies a small set
of IDAT files, and records the cross-reactive probe reference used during probe
filtering.

```{r preprocessing-local-inputs, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
preprocessing_inputs <- dnaEPICO:::exampleMinfiIdatInputsDnaEpico(n = 6L)

names(preprocessing_inputs)
preprocessing_inputs$tempDir
basename(preprocessing_inputs$phenoFile)
head(basename(list.files(preprocessing_inputs$idatFolder, full.names = TRUE)), 4)
basename(preprocessing_inputs$crossReactivePath)
```

`preprocessing_inputs` is a small list returned by the internal example helper.
`names(preprocessing_inputs)` shows its main elements: `tempDir` is the
temporary working directory, `idatFolder` contains the copied example IDAT
files, `phenoFile` is the phenotype table used by the function, `targets` is the
same phenotype information already loaded into memory, `arrayType` and
`annotationVersion` describe the array platform, and `crossReactivePath` points
to the cross-reactive probe reference used later during filtering.

```{r preprocessing-local, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
preprocessResult <- preprocessingMinfiEwasWater(
  phenoFile = preprocessing_inputs$phenoFile,
  idatFolder = preprocessing_inputs$idatFolder,
  outputLogs = file.path(preprocessing_inputs$tempDir, "logs"),
  nSamples = 6,
  SampleID = "Sample_Name",
  arrayType = preprocessing_inputs$arrayType,
  annotationVersion = preprocessing_inputs$annotationVersion,
  scriptLabel = "preprocessingMinfiEwasWater",
  baseDataFolder = file.path(preprocessing_inputs$tempDir, "rData"),
  figureBaseDir = file.path(preprocessing_inputs$tempDir, "figures"),
  detPThreshold = 1,
  normMethods = "quantile",
  sexColumn = "Sex",
  pvalThreshold = 1,
  chrToRemove = "",
  snpsToRemove = "SBE",
  mafThreshold = 1,
  crossReactivePath = preprocessing_inputs$crossReactivePath,
  plotGroupVar = "Sex",
  lcRef = "saliva",
  phenoOrder = "Sample_Name;Sex;Basename;Sentrix_ID;Sentrix_Position",
  lcPhenoDir = preprocessing_inputs$tempDir,
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(preprocessResult)
names(preprocessResult)
head(preprocessResult$targets[, c(
  "Sample_Name",
  "Sex",
  "Sentrix_ID",
  "Sentrix_Position"
)])
```

The returned object has class `dnaEPICO_preprocessingMinfiEwasWater`.
`names(preprocessResult)` lists the main elements returned by the function. The
`targets` element is the filtered phenotype table aligned to the retained
samples. The remaining elements are nested result objects for the major
preprocessing stages, including the `RGSet`, the methylation metrics, and the
cell-composition results.

Arguments:

- `phenoFile` and `idatFolder` define the phenotype table and the directory of
  raw IDAT pairs.
- `nSamples = 6` keeps the example small enough for interactive use.
- `SampleID = "Sample_Name"` identifies the column used to align phenotype data
  and methylation objects.
- `arrayType` and `annotationVersion` must match the array platform being
  processed.
- `detPThreshold`, `normMethods`, `pvalThreshold`, `chrToRemove`,
  `snpsToRemove`, and `mafThreshold` control quality filtering and probe
  filtering.
- `crossReactivePath` supplies the cross-reactive probe reference used during
  filtering.
- `sexColumn`, `plotGroupVar`, `lcRef`, and `phenoOrder` control sex
  prediction, quality-control grouping, cell-type estimation, and the structure
  of the returned phenotype table.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
  `saveOutputs = FALSE` keep the example quiet, in memory, and free of file
  outputs.

The next chunk inspects three important elements: the filtered `RGSet`, the
metric matrices, and the phenotype table augmented with cell-composition
estimates.

```{r preprocessing-local-inspect, eval = can_run_preprocessing, message = FALSE, warning = FALSE}
class(preprocessResult$RGSet)
names(preprocessResult$metricsData)
head(preprocessResult$lcData$phenoLC[, 1:6])
```

In local use, these are the most useful objects to inspect first.
`preprocessResult$RGSet` is the core methylation object. `metricsData`
contains the beta-value, M-value, and copy-number matrices used in downstream
steps. `lcData$phenoLC` is the phenotype table augmented with estimated cell
composition and is the object passed forward to later workflow stages.

# Step 2: svaEnmix

The SVA function loads a phenotype table and a saved `RGSet`, estimates
surrogate variables from ENmix control probes, and returns both the surrogate
variables and the merged phenotype table.

### Create example input files

The next chunk shows how to recreate the temporary files consumed by
`svaEnmix()`: a phenotype CSV file and a saved `RGSet` object in `.RData`
format.

```{r sva-local-inputs, eval = can_run_sva, message = FALSE, warning = FALSE}
sva_inputs <- dnaEPICO:::exampleMinfiBaseDataDnaEpico()
sva_temp_dir <- tempdir()
sva_targets_file <- file.path(sva_temp_dir, "sva_targets.csv")
sva_rgset_file <- file.path(sva_temp_dir, "sva_RGSet.RData")

names(sva_inputs)
class(sva_inputs$RGSet)
utils::write.csv(sva_inputs$targets, sva_targets_file, row.names = FALSE)
RGSet <- sva_inputs$RGSet
save(RGSet, file = sva_rgset_file)

basename(c(sva_targets_file, sva_rgset_file))
file.exists(c(sva_targets_file, sva_rgset_file))
```

`sva_inputs` is another helper list used to make the vignette self-contained.
`names(sva_inputs)` shows that it provides the in-memory `RGSet`, the matching
`targets` table, and a `crossReactivePath` reference. In this step, the function
needs file inputs rather than the in-memory objects, so the chunk writes
`targets` to `sva_targets.csv` and saves the `RGSet` object to
`sva_RGSet.RData`.

```{r sva-local, eval = can_run_sva, message = FALSE, warning = FALSE}
svaResult <- svaEnmix(
  phenoFile = sva_targets_file,
  rgsetData = sva_rgset_file,
  outputLogs = file.path(tempdir(), "logs"),
  SampleID = "Sample_Name",
  arrayType = "IlluminaHumanMethylation450k",
  annotationVersion = "ilmn12.hg19",
  SentrixIDColumn = "Sentrix_ID",
  SentrixPositionColumn = "Sentrix_Position",
  ctrlSvaPercVar = 0.90,
  ctrlSvaFlag = 1,
  scriptLabel = "svaEnmix",
  display = FALSE,
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(svaResult)
names(svaResult)
dim(svaResult$svaData$sva)
head(svaResult$mergedPheno[, 1:min(6, ncol(svaResult$mergedPheno))])
```

The returned object has class `dnaEPICO_svaEnmix`, and `names(svaResult)`
shows the main SVA components returned by the function. The `svaData$sva`
matrix contains one row per sample and one column per surrogate variable. The
`mergedPheno` table is often the most convenient object to inspect because it
combines the original phenotype data and the new SVA columns in one place.

Arguments:

- `phenoFile` points to the phenotype table that already includes the cell
  composition estimates from preprocessing.
- `rgsetData` points to the saved `RGSet` used to derive ENmix control-probe
  surrogate variables.
- `SampleID`, `SentrixIDColumn`, and `SentrixPositionColumn` identify the
  sample and array-position columns used in the SVA analysis.
- `ctrlSvaPercVar = 0.90` asks the procedure to keep enough control-derived
  surrogate variables to explain 90% of the control-probe variance.
- `ctrlSvaFlag = 1` enables the control-based SVA workflow.
- `display = FALSE`, `verbose = FALSE`, `logs = FALSE`, and
  `saveOutputs = FALSE` keep the example interactive and in-memory.

The association analysis between the surrogate variables and array-position
metadata is stored separately in `analysisData`.

```{r sva-local-inspect, eval = can_run_sva, message = FALSE, warning = FALSE}
class(svaResult$analysisData)
names(svaResult$analysisData)
```

This object contains the fitted models and ANOVA summaries used to assess how
the surrogate variables relate to Sentrix chip and position effects.

# Step 3: preprocessingPheno

The phenotype-preparation function aligns the phenotype table with the beta,
M-value, and copy-number matrices; splits the data by timepoint; constructs a
combined longitudinal object; and prepares Clock Foundation export tables.

### Create example input files

The next chunk shows how to recreate the temporary phenotype and matrix files
used in this example. The helper writes a phenotype table plus aligned beta,
M-value, and copy-number objects to a temporary directory.

```{r preprocessingPheno-local-inputs, message = FALSE, warning = FALSE}
pheno_inputs <- dnaEPICO:::examplePreprocessingPhenoStateDnaEpico()

names(pheno_inputs)
pheno_inputs$tempDir
basename(c(
  pheno_inputs$phenoPath,
  pheno_inputs$betaPath,
  pheno_inputs$mPath,
  pheno_inputs$cnPath
))
```

`pheno_inputs` is a list that bundles both the temporary files and the
corresponding in-memory objects used in this stage. `names(pheno_inputs)` shows
that it contains the temporary directory, the phenotype table, the three saved
matrix files (`phenoPath`, `betaPath`, `mPath`, and `cnPath`), and precomputed
helper objects such as `metricsData`, `timepointData`, `combinedData`, and
`clockFoundation`.

```{r preprocessingPheno-local, message = FALSE, warning = FALSE}
phenoResult <- preprocessingPheno(
  phenoFile = pheno_inputs$phenoPath,
  betaPath = pheno_inputs$betaPath,
  mPath = pheno_inputs$mPath,
  cnPath = pheno_inputs$cnPath,
  SampleID = "Sample_Name",
  timeVar = "Timepoint",
  timepoints = "1,2",
  combineTimepoints = "1,2",
  outputPheno = file.path(pheno_inputs$tempDir, "data", "preprocessingPheno"),
  outputRData = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "metrics"
  ),
  outputRDataMerge = file.path(
    pheno_inputs$tempDir,
    "rData",
    "preprocessingPheno",
    "mergeData"
  ),
  sexColumn = "Sex",
  outputLogs = file.path(pheno_inputs$tempDir, "logs"),
  outputDir = file.path(pheno_inputs$tempDir, "clockFoundation"),
  verbose = FALSE,
  logs = FALSE,
  saveOutputs = FALSE
)

class(phenoResult)
names(phenoResult)
names(phenoResult$combinedData)
head(phenoResult$combinedData$phenoBeta[, 1:6])
```

The returned object has class `dnaEPICO_preprocessingPheno`, and
`names(phenoResult)` shows the main returned components. The
`combinedData` element groups the combined longitudinal objects, and
`phenoResult$combinedData$phenoBeta` is the key downstream table for modeling.
It combines phenotype columns and CpG beta values in a single data frame that
can be passed to the GLM or GLMM functions.

Arguments:

- `phenoFile`, `betaPath`, `mPath`, and `cnPath` define the phenotype table and
  the three methylation matrices that are aligned in this step.
- `SampleID = "Sample_Name"` is the key used to match samples across the loaded
  objects.
- `timeVar = "Timepoint"`, `timepoints = "1,2"`, and
  `combineTimepoints = "1,2"` define the separate exports and the combined
  longitudinal object.
- `outputPheno`, `outputRData`, `outputRDataMerge`, and `outputDir` still need
  to be specified even when `saveOutputs = FALSE`, because they define the
  canonical output structure for the function.
- `sexColumn = "Sex"` identifies the sex variable used by the helper functions.
- `verbose = FALSE`, `logs = FALSE`, and `saveOutputs = FALSE` keep the example
  focused on the returned object rather than file writing.

The Clock Foundation export objects are returned in the same result, which
makes it possible to inspect the export-ready tables before deciding whether
they should be used in a reporting or export workflow.

```{r preprocessingPheno-local-inspect, message = FALSE, warning = FALSE}
names(phenoResult$clockFoundation)
head(phenoResult$clockFoundation$phenoCF)
```

The `clockFoundation` element contains the export-ready objects, and
`phenoResult$clockFoundation$phenoCF` is the phenotype table prepared for
[Clock Foundation](https://dnamage.clockfoundation.org/user) use.

# Summary

In interactive local use, the most important habit is to inspect the returned
objects after each step. In practice, the following checks are enough to
confirm that the workflow is behaving as expected:

- `class(result)` to confirm which function produced the object
- `names(result)` to see the main return elements
- `head()` on the phenotype-like tables
- `class()` or `dim()` on the methylation objects and matrices

The next vignette shows how to use `saveOutputs = TRUE` and the Makefile
workflow. The file-based vignette also describes the regression and
mixed-effects model structures used downstream, including how phenotype-specific
PRS terms are added through `prsMap`.

# Basics

Date the vignette was generated.

```{r, echo = FALSE}
Sys.time()
```

Wallclock time spent generating the vignette.

```{r, echo = FALSE}
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r, echo = FALSE}
sessionInfo()
```

## Asking for help

As package developers, we try to explain clearly how to use our packages and in
which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve, so it is critical to learn where to ask for help. We would like
to highlight the [Bioconductor support site](https://support.bioconductor.org/)
as the main resource for getting help. Please remember to use the `dnaEPICO`
tag and check the
[older posts](https://support.bioconductor.org/tag/dnaEPICO/). If you want to
receive help, please provide a small reproducible example and your session
information so the source of the problem can be tracked efficiently.

# References
